---
title: "AlphaSimRBee"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AlphaSimRBee}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# L0 functions 

nColonies; input colonies; return n

nCaste; input Colony or Colonies; return n
nQueen
nWorkers 
nDrones 
nVirginQueens
nFathers

isQueenMated; input Pop, Colony, or Colonies; return logical

extractQueensYOB; input colony; return numeric; TODO input Pop, Colony or Colonies

computeQueensAge; input Pop or colony; return numeric; TODO input Pop, Colony or Colonies

getId; input Colony or Colonies; return character

hasSplit; input Colony or Colonies; return logical

hasSwarmed; input Colony or Colonies; return logical

hasSuperseded; input Colony or Colonies; return logical

isProductive; input Colony or Colonies; return logical

simulateHoneyBeeGenomes

getCsdHaplo; input Pop, Colony or Colonies; return numeric
getCsdGeno

# L1 functions

getCaste; input Colony or Colonies; return Pop
getQueen
getFathers
getVirginQueens
getWorkers
getDrones

crateFounderDrones; input Pop; return Pop

createWorkers; input Colony; return Pop

createDrones; input Colony; return Pop

createVirginQueens; input Colony; return Pop

createDCA; input Colony or Colonies; return Pop

pullIndFromPop; input Pop; return list of 2 Pop

pullDroneGroupsFromDCA; input Pop; list of n Pop

pullCaste; input Colony or Colonies; return list of Colony and Pop
pullQueen
pullVirginQueens
pullWorkers
pullDrones

crossVirginQueen; input 2 Pop; return Pop

# L2 functions

createColony; input Pop; return Colony

setQueensYOB; input Pop or Colony; return Pop or Colony

addWorkers; input Colony, return Colony

addDrones; input Colony, return Colony

addVirginQueens; input Colony, return Colony

reQueenColony; input Colony, return Colony

buildUpColony; input Colony, return Colony

replaceWorkers; input Colony, return Colony
replaceDrones

removeQueen; input Colony, return Colony
removeVirginQueen
removeWorkers
removeDrones

resetEvents; input Colony, return Colony

crossColony; input Colony, return Colony

collapseColony; input Colony, return Colony

swarmColony; input Colony, return list of 2 Colony

supersedeColony; input Colony, return Colony

splitColony; input Colony; return list of 2 Colony

setPhenoColony; input Colony, return Colony

# L3 functions

createColonies; input NULL, Colony, or Colonies; return Colonies

addColonyToTheColonies; input Colones and Colony; return Colonies; TODO rename to addColony?
assignColonyToColonies

selectColonies; input Colonies; return Colonies

pullColonies; input Colonies; return list of 2 Colonies

removeColonies; input Colonies; return Colonies

createMultipleVirginColonies; input Pop; return Colonies

createMultipleMatedColonies; input Pop; return Colonies

buildUpColonies; input Colonies; return Colonies

replaceWorkersColonies; input Colonies; return Colonies

replaceDronesColonies; input Colonies; return Colonies

reQueenColonies; input Colonies; return Colonies

collapseColonies; input Colonies; return Colonies

supersedeColonies; input Colonies; return Colonies

swarmColonies; input Colonies; return list of 2 Colonies

splitColonies; input Colonies; return list of 2 Colonies

crossColonies; input Colonies; return Colonies

setPhenoColonies; input Colonies; return Colonies

```{r setup}
library(AlphaSimRBee)
```
