---
title: "AlphaSimRBee"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AlphaSimRBee}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# L0 functions 

nColonies; input colonies; return n

nCaste; input Colony or Colonies; return n
nQueen
nFathers
nWorkers 
nDrones 
nVirginQueens

isQueenMated; input Pop, Colony, or Colonies; return logical

getQueensYearOfBirgh & getQueensYOB; input Pop, Colony or Colonies; return numeric; TODO w/o s?
getQueensAge; TODO w/o s?

getId; input Colony or Colonies; return character
getLocation; input Colony or Colonies; return numeric or list of numeric

hasSplit; input Colony or Colonies; return logical
hasSwarmed
hasSuperseded
isProductive

simulateHoneyBeeGenomes TODO

getCsdHaplo; input Pop, Colony or Colonies; return matrix TODO
getCsdGeno

getIbdHaplo; input Pop; return matrix
getQtlHaplo
getQtlGeno
getSegSiteHaplo
getSegSiteGeno
getSnpHaplo
getSnpGeno

getCasteIbdHaplo; input Colony or Colonies; return matrix or list of matrix
getQueensIbdHaplo
getFathersIbdHaplo
getVirginQueensIbdHaplo
getWorkersIbdHaplo
getDronesIbdHaplo
getColonyIbdHaplo

getCasteQtlHaplo; input Colony or Colonies; return matrix or list of matrix
getQueensQtlHaplo
getFathersQtlHaplo
getVirginQueensQtlHaplo
getWorkersQtlHaplo
getDronesQtlHaplo
getColonyQtlHaplo

getCasteQtlGeno; input Colony or Colonies; return matrix or list of matrix
getQueensQtlGeno
getFathersQtlGeno
getVirginQueensQtlGeno
getWorkersQtlGeno
getDronesQtlGeno
getColonyQtlGeno

getCasteSegSiteHaplo; input Colony or Colonies; return matrix or list of matrix
getQueensSegSiteHaplo
getFathersSegSiteHaplo
getVirginQueensSegSiteHaplo
getWorkersSegSiteHaplo
getDronesSegSiteHaplo
getColonySegSiteHaplo TODO

getCasteSegSiteGeno; input Colony or Colonies; return matrix or list of matrix
getQueensSegSiteGeno
getFathersSegSiteGeno
getVirginQueensSegSiteGeno
getWorkersSegSiteGeno
getDronesSegSiteGeno
getColonySegSiteGeno TODO

getCasteSnpHaplo; input Colony or Colonies; return matrix or list of matrix
getQueensSnpHaplo
getFathersSnpHaplo
getVirginQueensSnpHaplo
getWorkersSnpHaplo
getDronesSnpHaplo
getColonySnpHaplo TODO

getCasteSnpGeno; input Colony or Colonies; return matrix or list of matrix
getQueensSnpGeno
getFathersSnpGeno
getVirginQueensSnpGeno
getWorkersSnpGeno
getDronesSnpGeno
getColonySnpGeno TODO

# L1 functions

getCaste; input Colony or Colonies; return Pop
getQueen
getFathers
getVirginQueens
getWorkers
getDrones

createFounderDrones; input Pop; return Pop; TODO merge with createDrones?

createWorkers; input Colony; return Pop; TODO Colonies?
createDrones
createVirginQueens

createDCA; input Colony or Colonies; return Pop

pullIndFromPop; input Pop; return list of 2 Pop

pullDroneGroupsFromDCA; input Pop; list of n Pop

pullCaste; input Colony or Colonies; return list of Pop and Colony
pullQueen
pullVirginQueens
pullWorkers
pullDrones

crossVirginQueen; input 2 Pop; return Pop

setLocation; input Colony or Colonies; return Colony or Colonies

setQueensYearOfBirth & setQueensYOB; input Pop, Colony, or Colonies; return Pop, Colony, or Colonies; TODO w/o s??

# L2 functions

createColony; input Pop; return Colony

addWorkers; input Colony, return Colony; TODO some have Colony in name, some don't!?
addDrones
addVirginQueens; TODO add external virgin queens?
reQueenColony; TODO rename to replaceQueen()?
buildUpColony; TODO use both replace* and add*? rename to buildUp()?
replaceWorkers
replaceDrones
removeQueen
removeVirginQueen
removeWorkers
removeDrones
resetEvents
crossColony; TODO rename to cross()?
collapseColony; TODO rename to collapse()?
supersedeColony; TODO rename to supersede()?
setPhenoColony; TODO rename to setColonyPheno()?

swarmColony; input Colony, return list of 2 Colony; TODO rename to swarm()?
splitColony; TODO rename to split() (clash with other R func(s))?

# L3 functions

createColonies; input NULL, Colony, or Colonies; return Colonies

addColonyToTheColonies; input Colones and Colony; return Colonies; TODO rename to addColony?
assignColonyToColonies

selectColonies; input Colonies; return Colonies
removeColonies

pullColonies; input Colonies; return list of 2 Colonies

createMultipleVirginColonies; input Pop; return Colonies; TODO remove mulitiple? How is this different to createColonies?
createMultipleMatedColonies

buildUpColonies; input Colonies; return Colonies; TODO rename to buildUp()?
replaceWorkersColonies; TODO rename to replaceWorkers()?
replaceDronesColonies; TODO rename to replaceDrones()?
reQueenColonies; TODO rename to replaceQueen()?
collapseColonies; TODO rename to collapse()?
supersedeColonies; TODO rename to supersede()?

swarmColonies; input Colonies; return list of 2 Colonies; TODO rename to swarm()?
splitColonies; TODO rename to split() (clash with other R func(s))?

crossColonies; input Colonies; return Colonies; TODO rename to cross()?
setPhenoColonies; TODO rename to setColonyPheno?

```{r setup}
library(AlphaSimRBee)
```
