---
title: "SIMplyBee - quantitative genetics"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIMplyBee - quantitative genetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(package = "SIMplyBee")
library(package = "ggplot2")
```

```{r, include = TRUE}
# Founder genomes
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
```

# Honey yield 

```{r, include = TRUE}
# Global simulation parameters
SP <- SimParamBee$new(founderGenomes)

# Quantitative genetic parameters - for a single trait with queen and worker genetic effects
meanP <- c(20, 0)
varA <- c(1, 1 / SP$nWorkers) # TODO: if SP$nWorkers is fun, use value here!
corA <- matrix(data = c( 1.0, -0.5, 
                        -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = 100, mean = meanP, var = varA, corA = corA)

varE <- c(3, 3 / SP$nWorkers)
# TODO: what is a reasonable environmental correlation between queen and worker effects?
corE <- matrix(data = c( 1.0, -0.3, 
                        -0.3,  1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE)
SP$setCorE(corE = corE)
```

```{r, include = TRUE}
# Base population virgin queens
basePop <- createVirginQueens(founderGenomes)
basePop@gv
basePop@pheno
```

```{r, include = TRUE}
# Base population drones
drones <- createDrones(x = basePop[1:5], nInd = 5)
drones@gv
drones@pheno
```

```{r, include = TRUE}
# Colony
queen <- crossVirginQueen(pop = basePop[6], drones = drones, 
                          nFathers = 5)
colony <- createColony(x = queen)
colony
colony@queen@gv
colony@queen@pheno
```

```{r, include = TRUE}
# Check if colony is productive
isProductive(colony)

# Build up colony to reach productive stage
colony <- buildUpColony(colony)
isProductive(colony)
```

```{r, include = TRUE}
colony@queen@gv
colony@queen@pheno

head(colony@workers@gv)
head(colony@workers@pheno)
```

```{r, include = TRUE}
# Collate genetic and phenotype values
df <- data.frame(id = colony@workers@id,
                 father = colony@workers@father,
                 gvQueenEff = colony@workers@gv[, 1],
                 gvWorkerEff = colony@workers@gv[, 2],
                 pvQueenEff =  colony@workers@pheno[, 1],
                 pvWorkerEff = colony@workers@pheno[, 2])
```

```{r include = TRUE}
# Covariation between queen and worker effects in workers
p <- ggplot(data = df, aes(x = gvQueenEff, y = gvWorkerEff)) +
  geom_point() +
  theme_classic()
print(p)
```

```{r include = TRUE}
# Variation in worker genetic values by patriline - for queen effects
p <- ggplot(data = df, aes(x = gvQueenEff, colour = father)) +
  geom_density() +
  theme_classic()
print(p)
```

```{r include = TRUE}
# Variation in worker genetic values by patriline - for worker effects
p <- ggplot(df, aes(x = gvWorkerEff, colour = father)) +
  geom_density() +
  theme_classic()
print(p)
```

```{r include = TRUE}
# Colony genetic and phenotype value
# TODO: can we just sum up the colony@workers@pheno[, 2] or shall we do something else?
# TODO: can we just add up colony@queen@pheno[, 1] and colony@workers@pheno[, 2] or is there any other relationship?
colonyYield <- function(colony) {
  matrix(colony@queen@pheno[, 1] + sum(colony@workers@pheno[, 2]))
}
colony <- setPhenoColony(colony, FUN = colonyYield)
colony@pheno
```

# Honey yield and Calmness

```{r, include = TRUE}
# Global simulation parameters
SP <- SimParamBee$new(founderGenomes)

# Quantitative genetic parameters - for two traits with queen and worker genetic effects
# TODO: what is the scale of calmness "values"?
meanP <- c(20, 0, 0, 0)
# TODO: revise covariances (I made them up!)
varA <- c(1, 1 / SP$nWorkers, 1, 1 / SP$nWorkers)
corA <- matrix(data = c( 1.0, -0.5, 0.0, 0.0, 
                        -0.5,  1.0, 0.0, 0.0,
                         0.0,  0.0, 1.0, 0.4,
                         0.0,  0.0, 0.4, 1.0), nrow = 4, byrow = TRUE)
SP$addTraitA(nQtlPerChr = 100, mean = meanP, var = varA, corA = corA)

# TODO: revise covariances (I made them up!)
varE <- c(3, 3 / SP$nWorkers, 3, 3 / SP$nWorkers)
corE <- matrix(data = c( 1.0, -0.3, 0.0, 0.0,
                        -0.3,  1.0, 0.0, 0.0,
                         0.0,  0.0, 1.0, 0.2,
                         0.0,  0.0, 0.2, 1.0), nrow = 4, byrow = TRUE)
SP$setVarE(varE = varE)
SP$setCorE(corE = corE)
```

```{r, include = TRUE}
basePop <- createVirginQueens(founderGenomes)
basePop@gv
basePop@pheno
```

```{r, include = TRUE}
drones <- createDrones(x = basePop[1:5], nInd = 5)
drones@gv
drones@pheno
```

```{r, include = TRUE}
queen <- crossVirginQueen(pop = basePop[6], drones = drones, 
                          nFathers = 5)
colony <- createColony(x = queen)
colony@queen@gv
colony@queen@pheno
```

```{r, include = TRUE}
colony <- buildUpColony(colony)
colony
```

```{r include = TRUE}
# TODO: does calmness influence honey yield in any way?
colonyPheno <- function(colony) {
  cbind(colony@queen@pheno[, 1] + sum(colony@workers@pheno[, 2]),
        colony@queen@pheno[, 3] + sum(colony@workers@pheno[, 4]))
}
colony <- setPhenoColony(colony, FUN = colonyPheno)
colony@pheno
```

# Strength and honey yield

Here we will show how to use `nWorkers()` function to genetically and environmentally influence the colony strength and with this honey yield.

```{r, include = TRUE}
# Global simulation parameters
SP <- SimParamBee$new(founderGenomes)

# Quantitative genetic parameters - for two traits with queen and worker genetic effects
meanP <- c(100, 20, 0)
# TODO: revise covariances (I made them up!)
varA <- c(100, 1, 1 / SP$nWorkers)
corA <- matrix(data = c(1.0,  0.0,  0.0,
                        0.0,  1.0, -0.5, 
                        0.0, -0.5,  1.0), nrow = 3, byrow = TRUE)
SP$addTraitA(nQtlPerChr = 100, mean = meanP, var = varA, corA = corA)

# TODO: revise covariances (I made them up!)
varE <- c(300, 3, 3 / SP$nWorkers)
corE <- matrix(data = c(1.0,  0.0,  0.0,
                        0.0,  1.0, -0.3,
                        0.0, -0.3,  1.0), nrow = 3, byrow = TRUE)
SP$setVarE(varE = varE)
SP$setCorE(corE = corE)
```

```{r, include = TRUE}
basePop <- createVirginQueens(founderGenomes)
basePop@gv
basePop@pheno
```

```{r, include = TRUE}
drones <- createDrones(x = basePop[1:5], nInd = 5)
drones@gv
drones@pheno
```

```{r, include = TRUE}
queen <- crossVirginQueen(pop = basePop[6], drones = drones, 
                          nFathers = 5)
colony <- createColony(x = queen)
colony@queen@gv
colony@queen@pheno
```

```{r, include = TRUE}
nWorkersFun <- function(colony) {
  rpois(n = 1, lambda = colony@queen@pheno[, 1])
}
colony <- buildUpColony(colony, nWorkers = nWorkersFun)
colony
```

```{r include = TRUE}
colonyYield <- function(colony) {
  matrix(colony@queen@pheno[, 2] + sum(colony@workers@pheno[, 3]))
}
colony <- setPhenoColony(colony, FUN = colonyYield)
colony@pheno
```

# Swarming and honey yield

TODO: show how we could genetically and environmentally influence if colony will swarm (can we do this?) and with this honey yield.

TODO: colonyYield should use isProductive - 0 when FALSE
