---
title: "SIMplyBee - Quantitative genetics"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIMplyBee - Quantitative genetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE
)
```

# Introduction

This vignette describes and demonstrates how SIMplyBee implements quantitative
genetics principles for honeybees. Specifically, it describes four different
examples where we simulate:

  1. Honey yield - a single colony trait,
  
  2. Honey yield and Calmness - two colony traits,
  
  3. Colony strength and Honey yield - two colony traits where one trait impacts
     the other one via the number of workers, and
  
  4. Swarming and Honey yield - two colony traits where one trait impacts the
     other one via a colony event (swarming).

```{r load_packages}
library(package = "SIMplyBee")
library(package = "ggplot2")
```

```{r founderGenomes}
# Founder genomes
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
```

# Honey yield

TODO: say more and show some equations

```{r SimParamBee}
# Global simulation parameters
SP <- SimParamBee$new(founderGenomes)

# Quantitative genetic parameters - for a single colony trait that is impacted
# by queen and worker genetic effects - each represented by a different trait
mean <- c(20, 0)
varA <- c(1, 1 / SP$nWorkers)
corA <- matrix(data = c( 1.0, -0.5, 
                        -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = 100, mean = mean, var = varA, corA = corA,
             name = c("queenTrait", "workersTrait"))

varE <- c(3, 3 / SP$nWorkers)
# TODO: what is a reasonable environmental correlation between queen and worker effects?
corE <- matrix(data = c(1.0, 0.3, 
                        0.3, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
```

```{r basePop_virgin_queens}
# Base population virgin queens
basePop <- createVirginQueens(founderGenomes, n = 20)
head(basePop@gv)
# TODO: We lose trait names in createVirginQueens() #404
#       https://github.com/HighlanderLab/SIMplyBee/issues/404
head(basePop@pheno)
hist(basePop@pheno[, "queenTrait"])
hist(basePop@pheno[, "workersTrait"])
```

TODO: Mention that drones also get these values - they just don't express them

```{r basePop_drones}
# Base population drones
drones <- createDrones(x = basePop[1:5], nInd = 3)
head(drones@pheno)
```

```{r create_colony}
# Colony
queen <- cross(x = basePop[6], drones = drones)
colony <- createColony(x = queen)
colony
colony@queen@pheno
```

```{r build_up_colony}
# Check if colony is productive
isProductive(colony)

# Build up colony to reach productive stage
colony <- buildUp(colony)
isProductive(colony)
```

```{r queen_and_workers_values}
head(colony@workers@pheno)
```

```{r data.frame}
# Collate genetic and phenotype values
df <- data.frame(id = colony@workers@id,
                 father = colony@workers@father,
                 gvQueenTrait = colony@workers@gv[, "queenTrait"],
                 gvWorkersTrait = colony@workers@gv[, "workersTrait"],
                 pvQueenTrait =  colony@workers@pheno[, "queenTrait"],
                 pvWorkersTrait = colony@workers@pheno[, "workersTrait"])
head(df)
```

```{r plot_queen_vs_worker_values}
# Covariation between queen and worker trait genetic values in workers
p <- ggplot(data = df, aes(x = gvQueenTrait, y = gvWorkersTrait)) +
  geom_point() +
  theme_classic()
print(p)
```

```{r distribution_by_patriline}
# Variation in worker genetic values by patriline - for queen trait
p <- ggplot(data = df, aes(x = gvQueenTrait, colour = father)) +
  geom_density() +
  theme_classic()
print(p)
```

```{r distribution_by_patriline_2}
# Variation in worker genetic values by patriline - for workers trait
p <- ggplot(df, aes(x = gvWorkersTrait, colour = father)) +
  geom_density() +
  theme_classic()
print(p)
```

```{r fathers_values}
# Variation in patriline genetic values
getFathers(colony)@gv
```

```{r colony_pheno} 
# Colony phenotype value
calcColonyPheno(colony, queenTrait = "queenTrait", workersTrait = "workersTrait")
help(calcColonyPheno)
help(calcColonyPhenoFromCaste)
```

# Honey yield and Calmness

TODO: say more and show some equations

```{r SimParamBee_2}
# Global simulation parameters
SP <- SimParamBee$new(founderGenomes)

# Quantitative genetic parameters - for two traits with queen and worker genetic effects
# TODO: what is the scale of calmness "values"?
meanP <- c(20, 0, 0, 0)
# TODO: revise covariances (I made them up!)
varA <- c(1, 1 / SP$nWorkers, 1, 1 / SP$nWorkers)
corA <- matrix(data = c( 1.0, -0.5, 0.0, 0.0, 
                        -0.5,  1.0, 0.0, 0.0,
                         0.0,  0.0, 1.0, 0.4,
                         0.0,  0.0, 0.4, 1.0), nrow = 4, byrow = TRUE)
SP$addTraitA(nQtlPerChr = 100, mean = meanP, var = varA, corA = corA,
             name = c("yieldQueenTrait", "yieldWorkersTrait",
                      "calmQueenTrait", "calmWorkersTrait"))

# TODO: revise covariances (I made them up!)
varE <- c(3, 3 / SP$nWorkers, 3, 3 / SP$nWorkers)
corE <- matrix(data = c(1.0, 0.3, 0.0, 0.0,
                        0.3, 1.0, 0.0, 0.0,
                        0.0, 0.0, 1.0, 0.2,
                        0.0, 0.0, 0.2, 1.0), nrow = 4, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
```

```{r base_pop_and_colony}
basePop <- createVirginQueens(founderGenomes)
drones <- createDrones(x = basePop[1:5], nInd = 3)
queen <- cross(x = basePop[6], drones = drones)
colony <- createColony(x = queen)
colony <- buildUp(colony)
colony
colony@queen@pheno
head(colony@workers@pheno)
```

```{r colony_pheno_2}
calcMyColonyPheno <- function(colony) {
  yield <- calcColonyPhenoFromCaste(colony,
                                    queenTrait = "yieldQueenTrait",
                                    workersTrait = "yieldWorkersTrait")
  calmness <- calcColonyPhenoFromCaste(colony,
                                       queenTrait = "calmQueenTrait",
                                       workersTrait = "calmWorkersTrait",
                                       checkProduction = FALSE)
  return(cbind(yield, calmness))
}
calcColonyPheno(colony, FUN = calcMyColonyPheno)
```

# Strength and honey yield

Here we will show how to use `nWorkers` function to genetically and
environmentally influence the colony strength and with this honey yield.

TODO: describe which trait is which (nWorkersColonyPhenotype has queenTrait = 1!)

TODO: say more and show some equations

say that this is just an example assuming that workers don't contribute to how many eggs a queen lays phenotypically!

```{r SimParamBee_3}
# Global simulation parameters
SP <- SimParamBee$new(founderGenomes)

# Quantitative genetic parameters
# - the first trait has only the queen effect
# - the second trait has both queen and workers effects
mean <- c(100, 20, 0)
# TODO: revise covariances (I made them up!)
varA <- c(100, 1, 1 / mean[1]) # mean[1] will be average nWorkers
corA <- matrix(data = c(1.0,  0.0,  0.0,
                        0.0,  1.0, -0.5, 
                        0.0, -0.5,  1.0), nrow = 3, byrow = TRUE)
SP$addTraitA(nQtlPerChr = 100, mean = mean, var = varA, corA = corA,
             name = c("strengthQueenTrait", "yieldQueenTrait", "yieldWorkersTrait"))

# TODO: revise covariances (I made them up!)
varE <- c(300, 3, 3 / SP$nWorkers)
corE <- matrix(data = c(1.0, 0.0, 0.0,
                        0.0, 1.0, 0.3,
                        0.0, 0.3, 1.0), nrow = 3, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
```

```{r base_pop_and_colony_2}
basePop <- createVirginQueens(founderGenomes)
drones <- createDrones(x = basePop[1:5], nInd = 3)
queen <- cross(x = basePop[6], drones = drones)
colony <- createColony(x = queen)
colony@queen@pheno
```

```{r colony_strength}
# Use already provided function in the package - read about it
# TODO: All functions using n*() functions should get ... so we can pass arguments to them
#       https://github.com/HighlanderLab/SIMplyBee/issues/405
colony <- buildUp(colony, nWorkers = nWorkersColonyPhenotype)
colony
help(nWorkersColonyPhenotype)
```

```{r colony_pheno_3}
calcMyColonyPheno <- function(colony) {
  yield <- calcColonyPhenoFromCaste(colony,
                                    queenTrait = "yieldQueenTrait",
                                    workersTrait = "yieldWorkersTrait")
  return(matrix(yield))
}
calcColonyPheno(colony, FUN = calcMyColonyPheno)
```

# Swarming and honey yield

TODO: show how we could genetically and environmentally influence if colony will swarm (can we do this?) and with this honey yield. https://github.com/HighlanderLab/SIMplyBee/issues/105

TODO: say more and show some equations

```{r SimParamBee_4}
# TODO
```
