---
title: "SIMplyBee - Crossing"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIMplyBee - Crossing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE
)
```
# Introduction

This vignette shows how you can cross virgin queens in SIMplyBee. Here, we present, how you can cross:
* single or multiple virgin queens (class `Pop`), virgin queen in a colony (class `Colony`), or all the virgin queens in an apiary or population (class `MultiColony`);
* cross either with pre-selected population/group of drones or according to a cross plan.
* cross queens on an open DCA or mating station

Here I define some functions for this vignette to work - until they are incorporated into SIMplyBee.
```{r}
#' @rdname createMatingStationDCA
#' @title Create a DCA of drones at a mating stations
#'
#' @description Level 1 function that creates a DCA at a classical honeybee
#'   mating station of several sister drone producing queens. The 
#'   functions first creates multiple drone producing queens (DPQs) from one colony; 
#'   and second, produces drones from the DPQs. All the created drones forms a 
#'   DCA at a mating station.
#'
#' @param colony \code{\link{Colony-class}} to produce drone producing queens from
#' @param nDPQs integer, the number of drone producing queens
#' @param nDronePerDPQ integer, number of drones each DPQ contributed to the DCA
#'
#' @return \code{\link{Pop-class}} with created drones resembling a DCA at a mating station
#'
#' @examples
#' founderGenomes <- quickHaplo(nInd = 10, nChr = 1, segSites = 100)
#' SP <- SimParamBee$new(founderGenomes)
#' basePop <- createVirginQueens(founderGenomes)
#' drones <- createDrones(basePop[1], n = 1000)
#' droneGroups <- pullDroneGroupsFromDCA(drones, n = 10, nDrones = 10)
#'
#' # Create a colony and cross it
#' colony1 <- createColony(x = basePop[2])
#' colony1 <- cross(colony1, drones = droneGroups[[1]])
#'
#' # Create a empty colony
#' colony2 <- createColony(x = basePop[3])
#' 
#' # Create a mating station from colony1
#' matingStation <- createMatingStation(colony1, nDPQs = 20, nDronesPerDPQ = 10)
#' 
#' # Cross colony2 on the mating station
#' fathers <- pullDroneGroupsFromDCA(matingStation, n = 1, nDrones = 15)
#' colony2 <- cross(colony2, drones = fathers) 
#' nFathers(colony2)
#' 
#' @export
createMatingStationDCA <- function(colony, nDPQs = 30, nDronePerDPQ = 100) {
  DPQs <- createVirginQueens(colony, nInd = nDPQs)
  drones <- createDrones(DPQs, nInd = nDronePerDPQ)
  return(drones)
}


#' @rdname createRandomCrossPlan
#' @title Create a cross plan by randomly sampling drones for each queen
#'   from a drone population
#'
#' @description Level 0 function that creates a cross plan by randomly sampling
#'   a desired number of drones from a DCA and assigning them to either
#'   virgin queen or colony ID
#'
#' @param IDs numeric, IDs of either the virgin queens OR the colonies (can't have both
#'   in the same cross plan!)
#' @param drones \code{\link{Pop-class}}, drone population available for mating (DCA)
#' @param nFathers integer or function, number of drones to be mated with each virgin
#'   queen
#'
#' @return named list with names being virgin queen or colony input IDs with each
#'   list element holding the IDs of selected drones
#'
#' @examples
#' founderGenomes <- quickHaplo(nInd = 15, nChr = 1, segSites = 100)
#' SP <- SimParamBee$new(founderGenomes)
#' basePop <- createVirginQueens(founderGenomes)
#' drones <- createDrones(basePop[1], n = 1000)
#' droneGroups <- pullDroneGroupsFromDCA(drones, n = 10, nDrones = 10)
#'
#' # Create an apiary of colonies and cross it
#' apiary <- createMultiColony(x = basePop[2:5])
#' apiary <- cross(apiary, drones = droneGroups[1:4])
#' apiary <- buildUp(apiary, nWorkers = 100, nDrones = 100)
#'
#' # Create a DCA from the drones from the apiary
#' DCA <- createDCA(apiary)
#' DCA # Inspect the DCA
#' 
#' # Create virgin queens, a virgin colony, and a virgin apiary
#' virginQueen1 <- basePop[6]
#' virginQueen2 <- basePop[7]
#' colony1 <- createColony(basePop[7])
#' apiary1 <- createMultiColony(basePop[8:10])
#' 
#' # Create a combined cross plan for mating the virgin queens (with virgin queen IDs)
#' crossPlanVirginQueens <- createRandomCrossPlan(IDs = c(virginQueen1@id, virginQueen2@id),
#'                                                        drones = DCA,
#'                                                        nFathers = nFathersPoisson)
#' 
#' # Cross the virgin queens according to the cross plan
#' virginQueen1 <- cross(virginQueen1, drones = DCA, crossPlan = crossPlanColonies)
#' virginQueen2 <- cross(virginQueen2, drones = DCA, crossPlan = crossPlanColonies)
#' 
#' # Create a cross plan for mating the virgin colonies and the virgin apiary (with colony IDs)
#' crossPlanColonies <- createRandomCrossPlan(IDs = getId(c(colony1, apiary1)),
#'                                            drones = DCA,
#'                                            nFathers = nFathersPoisson)
#' 
#' # Cross the colonies according to the cross plan
#' colony1 <- cross(colony1, drones = DCA, crossPlan = crossPlanColonies)
#' apiary1 <- cross(apiary1, drones = DCA, crossPlan = crossPlanColonies)
#' nFathers(colony1)
#' nFathers(apiary1)
#' 
#' # You can mate virgin queens and colonies on the same way on the mating stations's DCA
#' # Create a mating station from colony1
#' matingStationDCA <- createMatingStationDCA(colony1, nDPQs = 20, nDronePerDPQ = 10)
#' 
#' # Create another virgin apiary
#' apiary2 <- createMultiColony(basePop[11:13])
#' 
#' # Create a cross plan for crossing the apiary on the mating station
#' crossPlanApiary <- createRandomCrossPlan(IDs = getId(apiary2), drones = matingStationDCA, nFathers = nFathersPoisson)
#' 
#' # Cross the apiary
#' apiary2 <- cross(apiary2, drones = matingStationDCA, crossPlan = crossPlanApiary)
#' nFathers(apiary2)
#' 
#' @export
createRandomCrossPlan <- function(IDs, drones, nFathers) { # Do we want nFathers or nDrones???!!!
  if (!isPop(drones)) {
    stop("Argument drones must be a Pop class!")
  }
  drones <- drones[isDrone(drones)]
  if (is.function(nFathers)) {
    nFathers <- nFathers(length(IDs))
    fathersMatch <- sample(unlist(sapply(1:length(IDs),
                                         FUN = function(x) rep(x, nFathers[x]))))
  } else {
    fathersMatch <- sample(unlist(sapply(1:length(IDs),
                                         FUN = function(x) rep(x, nFathers))))
  }
  fatherIDs <- sample(as.vector(drones@id), size = length(fathersMatch), replace = F)
  crossPlan <- base::split(fatherIDs, fathersMatch)
  names(crossPlan) <- IDs
  return(crossPlan)
}


#' @rdname cross
#' @title Cross (mate) virgin queen(s) as a population, of a colony, or
#'   of all given colonies
#'
#' @description Level 1 function that crosses (mates) a virgin queen to a group
#'   of drones. The virgin queen(s) could be within a population (\code{\link{Pop-class}}),
#'   in a colony (\code{\link{Colony-class}}), or multi-colony (\code{\link{MultiColony-class}}).
#'   This function does not create any progeny, it only stores the
#'   mated drones (fathers) so we can later create progeny as needed.
#'   When input is a  (\code{\link{Colony-class}}) or  (\code{\link{MultiColony-class}}),
#'   one virgin queens is selected at random, mated, and promoted to the queen of the colony.
#'   Other virgin queens are destroyed. Mated drones (fathers) are stored for
#'   producing progeny at a later stage.
#'
#' @param x \code{\link{Pop-class}} or code{\link{Colony-class}} or \code{\link{MultiColony-class}},
#'   one or more virgin queens / colonies to be mated;
#' @param drones \code{\link{Pop-class}} or a list of \code{\link{Pop-class}},
#'   group(s) of drones that will be mated with virgin queen(s);
#'   if there is more than one virgin queen, the user has to provide
#'   a list of drone \code{\link{Pop-class}}. For this, the user can use
#'   \code{\link{pullDroneGroupsFromDCA}}
#' @param crossPlan, named list with names being virgin queen or colony input IDs with each
#'   list element holding the IDs of selected drones. Also see \code{\link{createRandomCrossPlan}}.
#7    If cross plan is NULL, we cross each virgin queens with the element-wise element
#'    of \code{drones}, which should be the sam length as the number of virgin queens
#'    If the cross plan is provided, the drones argument must be a single \code{\link{Pop-class}}.
#' @param removeFathers logical, removes those \code{drones} that have already
#'   mated; set to \code{FALSE} if you would like to mate a drone to multiple
#'   virgin queens, say via insemination
#' @param checkMating character, throw a warning (when \code{checkMating = "warning"}),
#'  or stop error (when \code{checkMating = "error"}) when some matings fail (see
#'  Details)
#' @param simParamBee \code{\link{SimParamBee}}, global simulation parameters
#'
#' @details This function changes caste for the mated drones to fathers, and
#'   mated virgin queens to queens. See examples. This means that you can not
#'   use these individuals in matings anymore!
#'
#' @seealso \code{\link{Colony-class}} on how we store the fathers along the
#'   queen.
#'
#' @details If the supplied drone population is empty (has 0 individuals), which
#'   can happen in edge cases or when \code{\link{nFathersPoisson}} is used
#'   instead of \code{\link{nFathersTruncPoisson}}, then mating of a virgin
#'   queen will fail and she will stay virgin. This can happen for just a few
#'   of many virgin queens, which can be annoying to track down, but you can use
#'   \code{\link{isQueen}} or \code{\link{isVirginQueen}}  to find such virgin queens. You can use
#'   \code{checkMating} to alert you about this situation.
#'
#' @return \code{\link{Pop-class}} with mated queen(s). The misc slot of the
#'   queens contains additional information about the number of workers, drones,
#'   and homozygous brood produced, and the expected percentage of csd homozygous
#'   brood.
#'
#' @examples
#' founderGenomes <- quickHaplo(nInd = 20, nChr = 1, segSites = 100)
#' SP <- SimParamBee$new(founderGenomes)
#' basePop <- createVirginQueens(founderGenomes)
#'
#' drones <- createDrones(x = basePop[1], nInd = 200)
#' droneGroups <- pullDroneGroupsFromDCA(drones, n = 8, nDrones = nFathersPoisson)
#'
#' # If input is a Pop class of virgin queen(s)
#' virginQueen1 <- basePop[2]
#' isQueen(virginQueen1)
#' (matedQueen1 <- cross(
#'   x = virginQueen1,
#'   drones = droneGroups[[1]]
#' ))
#'
#' isQueen(virginQueen1)
#' isQueen(virginQueen1)
#' isQueen(matedQueen1)
#' isQueen(matedQueen1)
#' nFathers(matedQueen1)
#' getFathers(matedQueen1)@id
#'
#' isDrone(getFathers(matedQueen1))
#' isFather(getFathers(matedQueen1))
#' isVirginQueen(matedQueen1)
#' isQueen(matedQueen1)
#'
#' virginQueen2 <- basePop[3]
#' (matedQueen2 <- cross(
#'   x = virginQueen2,
#'   drones = droneGroups[[2]]
#' ))
#' isQueen(virginQueen2)
#' isQueen(matedQueen2)
#' nFathers(matedQueen2)
#' getFathers(matedQueen2)@id
#'
#' matedQueens <- cross(
#'   x = c(basePop[4], basePop[5]),
#'   drones = droneGroups[c(3, 4)]
#' )
#' matedQueens
#' isQueen(matedQueens)
#' isQueen(matedQueens)
#' nFathers(matedQueens)
#' getFathers(matedQueens)
#'
#' # Inbred mated queen (mated with her own sons)
#' matedQueen3 <- cross(
#'   x = basePop[1],
#'   drones = droneGroups[[5]]
#' )
#' # Check the expected csd homozygosity
#' pHomBrood(matedQueen3)
#'
#' # If input is a Colony or MultiColony class
#' # Create Colony and MultiColony class
#' colony <- createColony(basePop[6])
#' isVirginQueen(getVirginQueens(colony))
#' apiary <- createMultiColony(basePop[7:8], n = 2)
#' all(isVirginQueen(mergePops(getVirginQueens(apiary))))
#'
#' # Cross
#' colony <- cross(colony, drones = droneGroups[[6]])
#' isQueenPresent(colony)
#' apiary <- cross(apiary, drones = droneGroups[c(7, 8)])
#' all(isQueenPresent(apiary))
#' nFathers(apiary)
#'
#' # Try mating with drones that were already used for mating
#' colony <- createColony(basePop[9])
#' try((matedColony <- cross(x = colony, drones = droneGroups[[1]])))
#' # Create new drones and mate the colony with them
#' drones <- createDrones(x = basePop[1], nInd = 15)
#' all(isDrone(drones))
#' any(isFather(drones))
#' (matedColony <- cross(x = colony, drones = drones))
#' isQueenPresent(matedColony)
#'
#' # For crossing virgin queens according to a cross plan, see
#'   \code{\link{createRandomCrossPlan}}}
#' 
#' # For crossing virgin queens on a mating stations, see
#'   \code{\link{createMatingStationDCA}}
#' 
#' @export
cross <- function(x, drones,
                  crossPlan = NULL,
                  checkMating = "error",
                  simParamBee = NULL) {
  if (is.null(simParamBee)) {
    simParamBee <- get(x = "SP", envir = .GlobalEnv)
  }
  if (!is.null(crossPlan) && !isPop(drones)) {
    stop("When cross plan is given, drones must be supplied as a single population!")
  }
  if (isPop(x)) {
    if (any(!isVirginQueen(x))) {
      stop("Individuals in pop must be virgin queens!")
    }
    if (is.null(crossPlan) && nInd(x) != length(drones)) { #check for list of father pops
      stop("Length of argument drones should match the number of virgin queens!")
    }
    if (!is.null(crossPlan) && !all(getId(x) %in% names(crossPlan))) { #Check for cross plan
      stop("Cross plan does not include all the virgin queens!")
    }
    nVirginQueen <- nInd(x)

    if (nVirginQueen == 1) {
      if (!isPop(drones)) {
        stop("When a single virgin queen is given, argument drones must be a Pop class object!")
      }
      if (!is.null(crossPlan)) {
        drones <- drones[crossPlan[[as.character(x@id)]]]
      }
      if (!all(isDrone(drones))) {
        stop("Individuals in drones must be drones!")
      }
      if (drones@nInd > 0) {
        x@misc[[1]]$fathers <- drones
        simParamBee$changeCaste(id = x@id, caste = "Q")
        simParamBee$changeCaste(id = drones@id, caste = "F")

        x <- setMisc(x = x, node = "nWorkers", value = 0)
        x <- setMisc(x = x, node = "nDrones", value = 0)
        x <- setMisc(x = x, node = "nHomBrood", value = 0)
        if (isCsdActive(simParamBee = simParamBee)) {
          val <- calcQueensPHomBrood(x = x)
        } else {
          val <- NA
        }
        x <- setMisc(x = x, node = "pHomBrood", value = val)
      } else if (drones@nInd == 0) {
        msg <- "Mating failed!"
        if (checkMating == "warning") {
          warning(msg)
        } else if (checkMating == "error") {
          stop(msg)
        }
      }
      ret <- x
    } else {
      ret <- list()
      for (virginQueen in seq_len(nVirginQueen)) {
        if (is.null(crossPlan)) {
          virginQueensDrones <- drones[[virginQueen]]
        } else {
          virginQueensDrones <- drones[crossPlan[[as.character(x[virginQueen]@id)]]]
        }
        ret[[virginQueen]] <- cross(x[virginQueen],
                                    drones = virginQueensDrones,
                                    checkMating = checkMating, simParamBee = simParamBee
        )
      }
      ret <- mergePops(ret)
    }
  } else if (isColony(x)) {
    if (isQueenPresent(x)) {
      stop("Queen already present in the colony!")
    }
    if (!isVirginQueensPresent(x)) {
      stop("No virgin queen(s) in the colony to cross!")
    }
    if (!is.null(crossPlan) && !(x@id %in% names(crossPlan))) {
      stop("Cross plan does not include the colony!")
    }
    virginQueen <- selectInd(x@virginQueens, nInd = 1, use = "rand")
    if (is.null(crossPlan)) {
      colonyDrones <- drones
    } else {
      colonyDrones <- drones[crossPlan[[as.character(x@id)]]]
    }
    queen <- cross(
      x = virginQueen,
      drones = colonyDrones,
      checkMating = checkMating,
      simParamBee = simParamBee
    )
    if (isQueen(queen)) {
      x <- reQueen(x, queen)
      x <- removeVirginQueens(x)
    }
    ret <- x
  } else if (isMultiColony(x)) {
    nCol <- nColonies(x)
    if (nCol == 0) {
      ret <- createMultiColony()
    } else {
      ret <- createMultiColony(n = nCol)
      if (is.null(crossPlan) && nCol != length(drones)) {
        stop("Length of argument drones should match the number of colonies!")
      }
      if (!is.null(crossPlan) && !all(getId(x) %in% names(crossPlan))) {
        stop("Cross plan does not include all the colonies!")
      }
      for (colony in seq_len(nCol)) {
        if (is.null(crossPlan)) {
          colonyDrones <- drones[[colony]]
        } else {
          colonyDrones <- drones[crossPlan[[as.character(x[[colony]]@id)]]]
        }
        ret[[colony]] <- cross(
          x = x[[colony]],
          drones = colonyDrones,
          checkMating = checkMating,
          simParamBee = simParamBee
        )
      }
    }
  }
  validObject(ret)
  return(ret)
}
```


First, we have to create a founder population and some virgin queen, virgin colonies, and virgin apiaries
that we will later cross.
```{r}
# Simulate 40 founder genomes
founderGenomes <- quickHaplo(nInd = 50, nChr = 1, segSites = 100)
# Set global population paramaters
SP <- SimParamBee$new(founderGenomes)
# Create a base population of 40 virgin queens
basePop <- createVirginQueens(founderGenomes)

# Prepare virgin queens, colonies and apiaries for crossing
virginQueen1 <- basePop[1]
virginQueen2 <- basePop[2]
virginQueen3 <- basePop[3]
virginQueens1 <- basePop[4:6]
virginQueens2 <- basePop[7:9]
virginQueens3 <- basePop[10:12]
colony1 <- createColony(basePop[13])
colony2 <- createColony(basePop[14])
colony3 <- createColony(basePop[15])
colony4 <- createColony(basePop[16])
apiary1 <- createMultiColony(basePop[17:21])
apiary2 <- createMultiColony(basePop[22:26])
apiary3 <- createMultiColony(basePop[27:31])
apiary4 <- createMultiColony(basePop[32:41])
```

We will now create a groups of droned from the remaining queens, that will represent a drone congregation area (DCA).
```{r}
# Create a DCA from the remaining virgin queens
DCA <- createDrones(basePop[42:50], nInd = 1000)
```

# Cross virgin queens on an open DCA
## Cross by pre-selecting drone populations
We will start by crossing our populations and colonies to a pre-selected populations of drones. We pre-select the groups by pulling a desired number of drone packages from a DCA with the function `pullDroneGroupsFromDCA()`. To this function we need to provide a groups of drones (`DCA`), how many groups you want to pull from the DCA (`n`), and how many drones per group do you want (`nDrones`). For `nDrones`, you can either specify an integer or a sampling function, which results in a different number of drones in each of the pulled groups. We provide some suggestions of the sampling functions:
* nFathersPoisson: samples the number of drones from a Poisson distribution with a default mean of 15 (the user can specif a different mean)
* nFathersTruncatedPoisson: samples the number of drones from a Poisson distribution with a default mean of 15 (the user can specif a different mean)
* you can specify your own sampling function.

We can pull the drone groups out separately for each crossing or pull them out all at once.
```{r}
# Pre-select drone (father) populations from a DCA
droneGroups <- pullDroneGroupsFromDCA(DCA, n = 20, nDrones = nFathersTruncPoisson)
```


Now, we can cross our virgin queens with drone packages.
```{r}
# A single virgin queen
virginQueen1 <- cross(virginQueen1, drones = droneGroups[[1]])
nFathers(virginQueen1)
# Multiple virgin queens
virginQueens1 <- cross(virginQueens1, drones = droneGroups[2:4])
nFathers(virginQueens1)
# A colony
colony1 <- cross(colony1, drones = droneGroups[[5]])
nFathers(colony1)
# An apiary
apiary1 <- cross(x = apiary1, drones = droneGroups[6:10])
nFathers(apiary1)
```

## Cross by preparing a cross plan
Providing packages / populations of drones can be memory consuming, especially when dealing with a large number of colonies. An alternative is to provide a cross plan with IDs of the virgin queens or colonies and drones, and a single drone population with all the drones listed in the cross plan. You can create a cross plan with the function `createRandomCrossPlan()`. This function creates a cross plan by randomly sampling a desired number of drones from a DCA and assigning them to either virgin queen ID or colony ID. When crossing a virgin queen in a colony, you have to provide the colony ID, since there could be multiple virgin queens within the colony and the random selection of one happens only within the `cross()` function. To create a cross plan you therefore have to provide the IDs of either the virgin queens or the colonies you want to cross (but not both in the same cross plan!!!), the drone population, and the number of drones you want to mate a particular virgin queen with. This can again be a fixed number or a sampling function. We can create a separate cross plan for each mating or create one combined cross plan for multiple matings (but can not have virgin queen's and colonies's ID in the cross plan at the same time!!!). Here, we again mate a single virgin queen, a population of virgin queens, a colony, and an apiary.
```{r}
# Create a combined cross for matingn a single queen (virginQueen2) and a population
# of virgin queen (virginQueens2)
crossPlanQueens <- createRandomCrossPlan(IDs = c(getId(virginQueen2),
                                                 getId(virginQueens2)),
                                         drones = DCA,
                                         nFathers = 15)
# Cross a aingle virgin queen
virginQueen2 <- cross(virginQueen2, drones = DCA, crossPlan = crossPlanQueens)
nFathers(virginQueen2)
# Cross multiple virgin queens
virginQueens2 <- cross(virginQueens2, drones = DCA, crossPlan = crossPlanQueens)
nFathers(virginQueens2)

crossPlanColonies <- createRandomCrossPlan(IDs = c(getId(colony2), getId(apiary2)),
                                           drones = DCA,
                                           nFathers = nFathersPoisson)
# Cross a single colony
colony2 <- cross(colony2, drones = DCA, crossPlan = crossPlanColonies)
nFathers(colony2)
# Cross an apiary
apiary2 <- cross(x = apiary2, drones = DCA, crossPlan = crossPlanColonies)
nFathers(apiary2)
```

# Cross virgin queens on a mating station
Mating virgin queens on a mating station is no different than mating them in an open DCA - the different in the DCA itself. In the case of open mating, the DCA consists of drones from multiple queens, all of which are usually unknown. In the case of a mating station, the DCA consists of drones all coming from a sister group of drone producing queens (DPQ), the queen of which is also known. This allows us to track the pedigree also on the paternal side.

First, you need to create a mating station DCA. For this, we have created a function `createMatingStationDCA()`, that takes a single sire colony (queen of the DPQs). From the sire colony, we first produce a desired number of sister DPQs, and next produce a desired number of drones per DPQ. The produced drones represent the mating station's DCA.
```{r}
# Create a DCA at a mating station from colony1
matingStationDCA <- createMatingStationDCA(colony1, nDPQs = 20, nDronePerDPQ = 1000)
```

Here, we will mate a single colony and a group of colonies (`Multicolony` object `apiary3`) on a mating station. We will again start by preparing a combined mating with colony IDs, according to which we will mate our queens.
```{r}
# Mate only an apiary
crossPlanMatingStation <- createRandomCrossPlan(IDs = c(getId(colony3), 
                                                         getId(apiary3)),
                                                drones = matingStationDCA,
                                                nFathers = nFathersTruncPoisson)

# Cross a colony
colony3 <- cross(colony3, crossPlan = crossPlanMatingStation, drones = matingStationDCA)
nFathers(colony3)
# Cross an apiary
apiary3 <- cross(apiary3, crossPlan = crossPlanMatingStation, drones = matingStationDCA)
nFathers(apiary3)
```

# Cross virgin queens with different methods
It could happen, that you have e.g. 10 virgin colonies in an apiary - you would want to inseminate one of the virgin queens artificially with a single drone, take three of them on a mating station, and mate the rest of them openly on a local DCA. Since cross plan is a named list, you can concatenate multiple cross plans into one. Let's mate the multicolony `apiary4` in such a manner.
```{r}
# Create a single drone for single drone insemination
singleDrone = createDrones(colony2, nInd = 1)
# Create a cross plan for crossinge some of the colonies in an open DCA, some with single drone, and some on a mating station
crossPlanApiary4 <- c(
  createRandomCrossPlan(IDs = getId(apiary4)[1], drones = singleDrone, nFathers = 1),
  createRandomCrossPlan(IDs = getId(apiary4)[2:6], drones = DCA, nFathers = nFathersTruncPoisson),
  createRandomCrossPlan(IDs = getId(apiary4)[7:10], drones = matingStationDCA, nFathers = nFathersTruncPoisson)
  )

apiary4 <- cross(apiary4, crossPlan = crossPlanApiary4, drones = c(singleDrone, DCA, matingStationDCA))
nFathers(apiary4)
```




