---
title: "Variance calculations between individual and colony level values"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VarianceCalculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 80
    canonical: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE
)
```

The idea of this work is:

1)  We want to do honey bee simulations with individual queens, drones, and
    workers

2)  To do the above, we require individual variances and covariances

3)  Literature reports only variances and covariances at the colony level

4)  The aim of this paper is to show how to use variances and covariances at the
    colony level to do simulations at the individual level.

# Introduction WORK IN PROGRESS

This document develops theory for the variance of genetic and phenotypic values
between colonies with the aim to understand how to fine tune a SIMplyBee
simulation where we need variance components for queen and workers effects at an
individual honeybee level. The challenge here is that in reality we do not
observe such variances, but only on a colony level.

For example, in the vignette we assumed that phenotypic variance on the
individual level is about 1/4 due to genetic variation and about 3/4 due to
environmental variation, but looking at colony genetic and phenotypic variances
the ratio was aobut 1/2!

TODO: What do we do about this? It feels like we need a full paper on all of
this;) :(

INTRODUCTION

TODO: Topic 1

TODO: cite this paper that does honeybee simulations, but without actually doing
honeybee colony stuff etc. Heritability in honeybees paper:
<https://agritrop.cirad.fr/599190/1/ece3.7389.pdf>

TOOD: cite some classic papers on variances in honeybees

TODO: cite recent papers from German and Dutch colleagues on variances in
honeybees

Brascamp and Bijma (2014): Methods to estimate breeding values in honey bees
<https://gsejournal.biomedcentral.com/articles/10.1186/s12711-014-0053-9>

Brascamp and Bijma (2019): A note on genetic parameters and accuracy of
estimated breeding values in honey bees
<https://gsejournal.biomedcentral.com/articles/10.1186/s12711-019-0510-6>

Andonov et al.: Modeling honey yield, defensive and swarming behaviors of
Italian honey bees (Apis mellifera ligustica) using linear-threshold approaches
<https://doi.org/10.1186/s12863-019-0776-2>

To make use of the material in this document you should run the quantitative
genetics vignette - the first example. Then you can continue here!

Having multiple colonies and their values, we can now return to the "about"
point mentioned at the start of this section, when we defined quantitative
genetic parameters. Recall that our starting quantitative genetic parameters for
the queen and workers effects were:

We will follow the example from Brascamp and Bijma (2019) they have:

$\sigma_{g_q}^2=0.5$ - this is their $\sigma_{A_q}^2=0.5$

$\sigma_{g_w}^2=1.0$ - this is their $\sigma_{A_w}^2=1.0$ among unrelated
honeybees

$\sigma_{g_w}^2=0.32$ - this is their $\sigma_{\bar{A}_w}^2=0.32$ among averages
of groups of honeybees?

$\sigma_{{g_q},{g_w}}=-0.35$

$\sigma_e^2=2.0$

$h^2=(a_base\sigma_{g,w}^2 + \sigma_{{g,w},{g,q}} \sigma_{g,q}^2)/\sigma_{p,c}^2=???$

$h_w^2=a_base\sigma_{g,w}^2/\sigma_{p,c}^2=0.13$

$h_q^2=\sigma_{g,q}^2/\sigma_{p,c}^2=0.20$

$T^2=(sigma_{g,w}^2 + 2\sigma_{{g,w},{g,q}} \sigma_{g,q}^2)/\sigma_{p,c}^2=0.32$

$T_w^2=\sigma_{g,w}^2/\sigma_{p,c}^2=0.41$

$T_q^2=\sigma_{g,q}^2/\sigma_{p,c}^2=0.20$

$r_g = cor(g_w, g_q) = -0.50$

TODO: equation 3 in Brascamp and Bijma (2019) is highly relevant to our work
here - that equation shows connection between the variance of the sum (or
average) of worker genetic values for worker effect AND variance of worker
genetic values for worker effect - THIS IS WHAT WE WORK WITH HERE SO WHAT IS
NOVEL IN OUR WORK? I think its about providing a tool and demonstrate it
clearly. Also, they work with average, while we work with the sum?

TODO: show sensitivity to the number of workers - the larger the number of
workers, the smaller the variance between workers

```{r quan_gen_param_revision}
# Trait means
mean

# Trait genetic variation - on a per honeybee level
varA
corA
(covA <- corA * outer(X = sqrt(varA), Y = sqrt(varA), FUN = "*"))

# Trait environmental variation - on a per honeybee level
varE
corE
(covE <- corE * outer(X = sqrt(varE), Y = sqrt(varE), FUN = "*"))

# Trait phenotypic variation - on a per honeybee level
(covP <- covA + covE)
(corP <- cov2cor(covP))
(varP <- diag(covP))

# Expected phenotypic variation - on a per colony level
# TODO: there is in fact more stuff happening here - see below! 
nW <- SP$nWorkers # TODO: this is not correct - we have queen-workers pairs and workers-workers (see below!)
# Var(z)=Var(x+y)=Var(x)+Var(sum(y))=Var(x)+Var(y_1)+Var(y_2)+...
# this is all wrong likely ...
covA[1, 1] + nW * covA[2, 2] + 2 * k * covA[1, 2] # variance can not be negative!
covP[1, 1] + nW * covP[2, 2] + 2 * k * covP[1, 2]

# Observed variation - on a per colony level
var(colonyGv)
var(colonyPheno)
var(colonyGv) / var(colonyPheno)
```

Hmm, we have much higher genetic and phenotypic variances in simulation than
based simply looking at queen and n\*workers variances, as well as much higher
ratio between the two than the initial value of \~1/4! Why? Let's see. For
genetic value of a colony $g_c$ we add up the queen's genetic value for the
queen effect $g_{q,q}$ and workers' genetic values for the workers effect
$\Sigma_{i=1}^{n_w}(g_{i,w})$:

$g_c = g_{q,q} + \Sigma_{i=1}^{n_w}(g_{i,w}).$

Expectation of colony genetic value is then:

$E(g_c) = E(g_{q,q} + \Sigma_{i=1}^{n_w}(g_{i,w}))$

$E(g_c) = E(g_{q,q}) + E(\Sigma_{i=1}^{n_w}(g_{i,w}))$

$E(g_c) = \mu_{g_{q}} + n_w E(g_{w,w})$

$E(g_c) = \mu_{g_{q}} + n_w \mu_{g_{w}}$

we are assuming here that $n_w$ is a fixed value, but if it is a random variable
then we would have

$E(g_c) = \mu_{g_{q}} + E(n_w) E(g_{w,w})$ using
<https://en.wikipedia.org/wiki/Distribution_of_the_product_of_two_random_variables>

$E(g_c) = \mu_{g_{q}} + \lambda_{n_w} \mu_{g_{w}}$

where $\lambda_{n_w}$ is average number of workers. So, in our case this turns
out the same as above. So in our case we have:

```{r}
nW <- SP$nWorkers
mean[1] + nW * mean[2]
mean(colonyGv)
mean(colonyPheno)
```

Variance of colony genetic value is then:

$Var(g_c) = Var(g_{q,q} + \Sigma_{i=1}^{n_w}(g_{i,w}))$

$Var(g_c) = Var(g_{q,q}) + Var(\Sigma_{i=1}^{n_w}(g_{i,w})) + 2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w}))$

$Var(g_c) = \sigma^2_{g_{q}} + Var(\Sigma_{i=1}^{n_w}(g_{i,w})) + 2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w}))$

So, we have three parts:

A)  the variance of the queen effect in queens $\sigma^2_{g_{q}}$,

B)  variance of a sum of workers effect in workers
    $Var(\Sigma_{i=1}^{n_w}(g_{i,w}))$, and

C)  covariance between the two $2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w}))$.

# A) Variance of the queen effect in queens $\sigma^2_{g_{q}}$ - in our case this is:

```{r check_queen_variances}
varA[1]
var(getQueenGv(apiary, collapse = TRUE)[, "queenTrait"])
# ... note that R's var() divides by n-1, which matters with small n
sum((getQueenGv(apiary, collapse = TRUE)[, "queenTrait"] - mean[1])^2) / nColonies(apiary)
```

Anyway, quite close!

# B) Variance of a sum of workers effect in workers $Var(\Sigma_{i=1}^{n_w}(g_{i,w}))$

Given that this is a sum, we need to look at:

B1) variance of each of worker genetic values $Var(g_{i,w})$ and

B2) covariance between each pair of the values $Cov(g_{i,w}, g_{j,w})$.

This means, that family structure will start to matter because it induces
covariance between family members (workers). The simplest case is when all
workers come from the same father. Then we have:

$g_{1,w} = \frac{1}{2}g_{q,w} + g_{f,w} + r_{1,w}$

$g_{2,w} = \frac{1}{2}g_{q,w} + g_{f,w} + r_{2,w}$

...

$g_{n_w,w} = \frac{1}{2}g_{q,w} + g_{f,w} + r_{n_w,w}$

## B1) variance of each of worker genetic values $Var(g_{i,w})$

As stated initially, we have $Var(g_{i,w}) = \sigma^2_{g_{w}}$ and we have $n_w$
such terms, $n_w \sigma^2_{g_{w}}$. In our case this would be:

```{r check_workers_variances}
nW * varA[2]
g <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = "workersTrait")
var(g)
# ... note that R's var() divides by n-1, which matters with small n
sum((g - nW * mean[2])^2) / nColonies(apiary)
```

So, a huge discrepancy between the variance of a sum of genetic values for the
workers effect `nW * varA[2]` (assuming independence) and a realised sum of
genetic values for the workers effect. As we have seen above, this is due to
family structure and associated B2) covariance between each pair of the values
$Cov(g_{i,w}, g_{j,w})$.

## B2) covariance between each pair of the values $Cov(g_{i,w}, g_{j,w})$

Clearly, these covariances matter a lot - there is lots of worker pairs in a
colony! How do these covariances look like?

$Cov(g_{1,w}, g_{2,w}) = Cov(\frac{1}{2}g_{q,w} + g_{f,w}, \frac{1}{2}g_{q,w} + g_{f,w})$

$Cov(g_{1,w}, g_{2,w}) = Var(\frac{1}{2}g_{q,w}) + Var(g_{f,w})$

assuming that mother and father are not related; further genetic variance
between drones is in fact half the genetic variance between queens because they
are haploid, so we get:

$Cov(g_{1,w}, g_{2,w}) = \frac{1}{4}\sigma^2_{g_w} + \frac{1}{4}\sigma^2_{g_w}$

$Cov(g_{1,w}, g_{2,w}) = \frac{1}{2}\sigma^2_{g_w}$.

Every pair of super-sister workers adds $\frac{1}{2}\sigma^2_{g_w}$ to B - note
that pair A-B adds this value, but so does the pair B-A, hence total is
$2\frac{1}{2}\sigma^2_{g_w} = \sigma^2_{g_w}$. With $n_w$ workers we get
$n_w n_w$ pairs (including with itself) or $n_w n_w - n_w$ pairs between
different workers. The total covariance contribution is then
$(n_w n_w - n_w)\frac{1}{2}\sigma^2_{g_w}$). In our case this would be:

```{r check_workers_variances2}
(varSumWorkers <- nW * varA[2]) # B1
(covSumWorkers <- (nW * nW - nW) * 1/2 * varA[2]) # B2
varSumWorkers + covSumWorkers # B

g <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = "workersTrait")
var(g) # B
# ... note that R's var() divides by n-1, which matters with small n
sum((g - nW * mean[2])^2) / nColonies(apiary) # B
```

If we would only have super-sisters we would have \~1 for B1, \~49.5 for B2, and
\~50.5 for B, but with a mix of half-sisters and super-sisters we would have a
lower value, which is what we see above, \~26.8 for B.

TODO: Define the setting above in intro or start of M&M - We also can have
full-sisters, where the mother is obviously the same, but the fathers are
different, yet they come from the same mother, which is equivalent to a
full-sibs case in diploid species. We will assume in this work that we have
unrelated queens and unrelated drones (=base pop) - our aim is to set variances
in such a base population - to get the scale or the variances right. In
simulation of our base population we generated drones from virgin queens, which
means that our drones can in fact be brothers, so our realised variances might
be a bit off compared to this theory. But this is fine, we are mostly trying to
get the order of variances correct.

So, now we need to work out B2 with a mix of half-sisters and super-sisters :(
Sister workers are related only due to having the same mother, assuming that
drones the queen mated with are unrelated:

$g_{1,w} = \frac{1}{2}g_{q,w} + g_{f1,w} + r_{1,w}$

$g_{2,w} = \frac{1}{2}g_{q,w} + g_{f2,w} + r_{2,w}$

$Cov(g_{1,w}, g_{2,w}) = Cov(\frac{1}{2}g_{q,w} + g_{f1,w}, \frac{1}{2}g_{q,w} + g_{f2,w})$

$Cov(g_{1,w}, g_{2,w}) = Var(\frac{1}{2}g_{q,w})$

$Cov(g_{1,w}, g_{2,w}) = \frac{1}{4}Var(g_{q,w})$

Hence every pair of half-sister workers adds $\frac{1}{4}\sigma^2_{g_w}$ to B
(note, this is $2\frac{1}{4}\sigma^2_{g_w}$ for A-B and B-A pairs!). With $n_w$
workers and $n_f$ fathers we have $\frac{n_w}{n_f}$ workers per father. Further,
we have $n_f$ groups of super-sisters and $n_f n_f - n_f$ pairs of sister
groups. Assuming that half-sister groups are the same size, we have
$n_f \frac{n_w}{n_f} \frac{n_w}{n_f} = n_w \frac{n_w}{n_f} = \frac{n^2_w}{n_f}$
pairs of super-sisters and $(n_f n_f - n_f) \frac{n_w}{n_f} \frac{n_w}{n_f}$ -
this is
$n_f (n_f - 1) \frac{n_w}{n_f} \frac{n_w}{n_f} = (n_f - 1) n_w \frac{n_w}{n_f} = \frac{ (n_f - 1) n^2_w}{n_f}$.
Is this correct?

```{r pairs_of_workers}
(nF <- nFathers(colony))
nW * nW
(a <- nW * nW / nF)
(b <- (nF - 1) * nW * nW / nF)
a+b 
```

Looks like!

So, super-sister pairs add $\frac{n^2_w}{n_f} \frac{1}{2}\sigma^2_{g_w}$ and
half-sister pairs add $\frac{ (n_f - 1) n^2_w}{n_f} \frac{1}{4}\sigma^2_{g_w}$.
Let's test this:

```{r check_workers_variances3}
(varSumWorkers <- nW * varA[2]) # B1
(covSumWorkersSuperSisters <- (nW * nW / nF) * 0.50 * varA[2]) # B2 in super-sisters
(covSumWorkersHalfSisters <- ((nF - 1) * nW * nW / nF) * 0.25 * varA[2]) # B2 in half-sisters
varSumWorkers + covSumWorkersSuperSisters + covSumWorkersHalfSisters # B

g <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = "workersTrait")
var(g) # B
# ... note that R's var() divides by n-1, which matters with small n
sum((g - SP$nWorkers * mean[2])^2) / nColonies(apiary) # B
```

Cool - very close!!! There can be some difference because the formulae above
assume that we have fixed sizes of half-sister and super-sister groups. Also,
there can be full-sisters in there as well!!!

TODO: We could also add full-sisters: To do this, we would have to assume how
many brother drones we use to figure out the grups

TODO: What if $n_w$ is a random variable? Following
<https://en.wikipedia.org/wiki/Distribution_of_the_product_of_two_random_variables>
we would use:

$Var(XY) = (Var(X) + E(X)^2) (Var(Y) + E(Y)^2) - E(X)^2 E(Y)^2$

but this is for two variables, while I have $n_w$ genetic values, so this
reference result is not that useful in our case! Well, if $n_w$ varies between
colonies, this must boost variance of colony-level genetic values significantly,
because kind of start calculating variance between apples (small colonies) and
oranges (large colonies).

# C) Covariance between the two $2Cov(g_{q,q}, \Sigma_{i-1}^{n_w}(g_{i,w}))$.

Let's repeat:

$g_{1,w} = \frac{1}{2}g_{q,w} + g_{f1,w} + r_{1,w}$

$g_{2,w} = \frac{1}{2}g_{q,w} + g_{f2,w} + r_{2,w}$ ...

The essential bit here is:

$Cov(g_{q,q}, g_{i,w}) = Cov(g_{q,q}, \frac{1}{2} g_{q,w})$

assuming that the queen and fathers are unrelated. Then:

$Cov(g_{q,q}, g_{i,w}) = Cov(g_{q}, \frac{1}{2} g_{w})$

$Cov(g_{q,q}, g_{i,w}) = \frac{1}{2} Cov(g_{q}, g_{w})$

$Cov(g_{q,q}, g_{i,w}) = \frac{1}{2} \sigma_{g_{q},g_{w}}$

So:

$2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w})) = 2Cov(g_{q,q}, n_w \frac{1}{2} g_{q,w})$

$2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w})) = 2 n_w \frac{1}{2}Cov(g_{q,q}, g_{q,w})$

$2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w})) = n_w \sigma_{g_{q},g_{w}}$

```{r check_workers_variances4}
(varQueen <- varA[1]) # A

(varSumWorkers <- nW * varA[2]) # B1
(covSumWorkersSuperSisters <- (nW * nW / nF) * 0.50 * varA[2]) # B2 in super-sisters
(covSumWorkersHalfSisters <- ((nF - 1) * nW * nW / nF) * 0.25 * varA[2]) # B2 in half-sisters
(varSumWorkers <- varSumWorkers + covSumWorkersSuperSisters + covSumWorkersHalfSisters) # B

(covQueenSumWorkers <- nW * covA[1, 2]) # C

varQueen + varSumWorkers + covQueenSumWorkers

g <- colonyGv
var(g)
sum((g - 20)^2) / nColonies(apiary)


(varQueen <- varP[1]) # A

(varSumWorkers <- nW * varP[2]) # B1
(covSumWorkersSuperSisters <- (nW * nW / nF) * 0.50 * varA[2]) # B2 in super-sisters - only genetic cov here, I think TODO!?
(covSumWorkersSisters <- ((nF - 1) * nW * nW / nF) * 0.25 * varA[2]) # B2 in sisters - only genetic cov here, I think TODO!?
(varSumWorkers <- varSumWorkers + covSumWorkersSuperSisters + covSumWorkersSisters) # B

(covQueenSumWorkers <- nW * covP[1, 2]) # C - pheno or genetic only here - TODO!?

varQueen + varSumWorkers + covQueenSumWorkers

p <- colonyPheno
var(p)
sum((p - 20)^2) / nColonies(apiary)
```

Bingo for genetic variance and very close for phenotypic too, but we need to
check some part above!

# What now?

Alright, what can we now do with this? We now have a system of expected genetic
and phenotypic variances for colony values as a function of number of workers,
fathers, and genetic and phenotypic covariances between the queen and worker
effects. So, it should be possible, in principle, to solve for what kind of
covariance values for the queen and workers effects should we use, to get the
desired colony genetic and phenotypic variances!

TODO: develop such estimating equations!

So, we have genetic part:

$A_g = \sigma^2_{g_q}$

$B1_g = n_w \sigma^2_{g_w}$

$B2_g = \frac{n^2_w}{n_f} \frac{1}{2} \sigma^2_{g_w}$

$B3_g = \frac{(n_f - 1) n^2_w}{n_f} \frac{1}{4} \sigma^2_{g_w}$

$C_g = n_w \sigma^2_{{g_q},{g_w}}$

here are 3 unknowns: $\sigma^2_{g_q}$, $\sigma^2_{g_w}$, and
$\sigma^2_{{g_q},{g_w}}$

we can add phenotypic part:

$A_p = \sigma^2_{p_q}$

$B1_p = n_w \sigma^2_{p_w}$

$B2_p = \frac{n^2_w}{n_f} \frac{1}{2} \sigma^2_{g_w}$ TODO: check if its only
genetic

$B3_p = \frac{(n_f - 1) n^2_w}{n_f} \frac{1}{4} \sigma^2_{g_w}$ TODO: check if
its only genetic

$C_p = n_w \sigma^2_{{p_q},{p_w}}$

here are additional 3 unknowns: $\sigma^2_{p_q}$, $\sigma^2_{p_w}$, and
$\sigma^2_{{p_q},{p_w}}$

while we could provide as guidance 2, maybe 3, inputs: $\sigma^2_{g_c}$,
$\sigma^2_{p_c}$, and $\frac{\sigma^2_{g_c}}{\sigma^2_{p_c}}$.

TODO: are there any additional inputs that we could get from literature? How do
we connect all this with variance estimates for the queen and the workers effect
in the honeybee genetics and breeding literature? I think that those estimates
are essentially this:
$\sigma^2_{g_c} = \sigma^2_{g_{c,q}} + \sigma^2_{g_{c,w}} + 2\sigma_{g_{c,q},g_{c,w}}$
where $\sigma^2_{g_{c,q}}$ is the queen part of $\sigma^2_{g_c}$, but not
necessarily the same as $\sigma^2_{g_w}$, or is it? Maybe that is the same, and
$\sigma^2_{g_{c,w}}$ is a counterpart to
$n_w \sigma^2_{g_w} + \frac{n^2_w}{n_f} \frac{1}{2} \sigma^2_{g_w}, \frac{(n_f - 1) n^2_w}{n_f} \frac{1}{4} \sigma^2_{g_w}$
(TODO: should we add full-sisters?), while $\sigma_{g_{c,q},g_{c,w}}$ is a
counterpart to $n_w \sigma_{{g_q},{g_w}}$. If this is so, then it fells that we
should have 3 more inputs for genetic part, so 4 inputs to estimate 3 unknowns
of which some are very simple transformations!!!

Additional input could be $\sigma^2_{p_c}$. For sure we don't get estimates of
$\sigma^2_{p_c} = \sigma^2_{p_{c,q}} + \sigma^2_{p_{c,w}} + 2\sigma_{p_{c,q},p_{c,w}}$,
but, having sorted the genetic part and if $B2_p$ and $B3_p$ depend only on
genetic covariance, then maybe, we have some maneuvering space to estimate also
$\sigma^2_{p_q}$, $\sigma^2_{p_w}$, and $\sigma^2_{{p_q},{p_w}}$?!

Furthermore, we have some constraints, so this is not just any system of
equations. For example, we know that variances must be positive. Futhermore, we
might know that come covariances are either positive or negative.

TODO: Develop and polish all this further and create a Shiny application where
we can change these individual-level parameters and study how that will change
colony-level parameters and then how we could go the other way around based on
literature and solving the above systems of equations. We could host such a
Shiny app on SIMplyBee.info!? Can we?

