---
title: "SIMplyBee - Colony events"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIMplyBee - Colony events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE
)
```

# Introduction
This vignette will introduce you to colony events. We implemented most important 
biological peculiarities of honey bees like swarming, supersedure and collapse of the colony.
Beside that we included also beekeeping practice called splitting. Some of this of
this events influnce also genetics in the honey bee colony.

# Swarming
Swarming is the process in which honey bee colonies reproduce to form new colonies. 
When a honey bee colony outgrows its hive, becomes too congested, or too populated 
for the queen’s pheromones to spread among workers, then the swarming begins.  
The workers start building swarm cells for new queens. 
Once the swarm cells are constructed, and the queen lays eggs in them, then the 
colony changes its behavior. Foraging slows down, and the workers begin erratic 
movements within the hive. Meanwhile, the queen stops laying eggs and reduces her 
weight to be able to fly. When the queen is ready, she leaves the hive and is followed
by about half of the workers in a massive cloud of flying bees. The queen will 
find a nearby tree, land, and emit pheromones that signal the workers to cluster
around her. The cluster will remain there for several hours while scout bees explore 
the area for a new home. Often the cluster will leave, travel a mile or more, 
and reform on another branch far away from the original hive. The cluster will 
remain until scout bees locate a new hive or cavity to occupy. Once a suitable
home is found, the entire cluster will take flight and move to the new location 
where it will begin building comb, developing new brood, and gathering pollen and nectar.

The queens developing in the swarm cells in the old hive are attended by half of 
the workers that did not leave with the swarm. After 16 days, the new queens begin 
to emerge. Typically, the first queen to emerge will sting the other queens in 
their cells and kill them in order to assume the role as the new queen for the 
colony. She then will take flight to find drones to mate with in order 
to begin the process of egg-laying and rebuilding the workforce in the colony. 
(source: https://hgic.clemson.edu/factsheet/frequently-asked-questions-about-honey-bee-swarms/#:~:text=What%20Is%20A%20Swarm%20Of,it%20is%20time%20to%20swarm.). The queen in remnant is daughter of a queen that swarmed. 

Our function for swarming splits the number of workers in a colony according to specified 
argument p (percentage of swarming). The function works both on colony and multcolony objects. 
Return of the function is a list with two Colony or MultiColony classes, swarm and remnant.
Remnant contains the old queen and specified percentage of workers. the remnant contains the rest of 
workers, all the drones and virgin queen that is daughter of a queen that swarmed.
TODOD: describe how to elect specified %of colonies to swarm
```{r create founder genoms and a Colony and Multicolony objects }
founderGenomes <- quickHaplo(nInd = 10, nChr = 1, segSites = 100)
SP <- SimParamBee$new(founderGenomes)
basePop <- createVirginQueens(founderGenomes)
drones <- createDrones(basePop[1], n = 1000)
fatherGroups <- pullDroneGroupsFromDCA(drones, n = 10, nFathers = 10)

# Create Colony and MultiColony class
colony <- createColony(x = basePop[2])
colony <- cross(colony, fathers = fatherGroups[[1]])
(colony <- buildUp(colony, nWorkers = 100))
apiary <- createMultiColony(basePop[3:8], n = 6)
apiary <- cross(apiary, fathers = fatherGroups[2:7])
apiary <- buildUp(apiary, nWorkers = 100)
```
TODO: finish example
# Supersedure
Supersedure is a replacement of a queen by her daughter without interference of beekeeper.
Supersedure is a natural way of requeening without swarming. there are many reasons for
supersedure: poor physical condition of a queen, old age, diseases, depleted spermatheca, 
poorly bread queen, reduced pheromone output and many others (source: https://www.tandfonline.com/doi/pdf/10.1080/0005772X.2010.11417360#:~:text=Supersedure%2C%20as%20it%20relates%20to,summer%20or%20in%20early%20autumn.). 
Our function supersede works both on colony and Multicolony objects. nVirginQueens parameter specifies 
the number of virgin queens of which one would be randomly selected for queen.
New queen is not mated yet.
# Collapse
Collapse of the colony is a term that describes death of a colony. all individuals 
within a colony die when collapse occurs. There are many reasons for collapse of a honey bee colony like:
diseases, starvation, queen problems, environmental and in-hive contamination with pesticides.
Colony losses can be very high, up to 60% of colonies per year. High colony losses can significantly influence
genetic structure of a population. 
Our function works both on Colony and Multicolony objects. 
# Splitting 
Colony splitting is a common beekeeping technique for managing swarming. 
A percentage of workers, brood and food stores are removed and combined in a new colony. 
Old queen normally stays with splited colony (remnant in our function). Function 
works on Colony and Multicolonyobjects. Argument p is proportion of workers that will go to 
new colony.

TODO: add colony events images!

TODO: where do we describe L0, L1, L2, and L3? In the next vignette that dives
      into these aspects (already?)?
TODO: related to the above, where do we want to write a list of functions
      grouped by some nice system (according to L0, L1, L2 and L3?)


TODO: add colony events images!

TODO: where do we describe L0, L1, L2, and L3? In the next vignette that dives
      into these aspects (already?)?
TODO: related to the above, where do we want to write a list of functions
      grouped by some nice system (according to L0, L1, L2 and L3?)
