---
title: "SIMplyBee - Colony events"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SIMplyBee - Colony events}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE
)
```

# Introduction

This vignette will introduce you to colony events. We implemented the most important natural events like swarming, supersedure, and collapse
of the colony, and a beekeeping management practice called splitting. All functions that implement colony events work both on `Colony` and `MultyColony` objects.

# Swarming

Swarming is the process in which honeybee colonies reproduce to form new colonies. When a honeybee colony outgrows its hive, becomes too
congested, or too populated for the queen's pheromones to spread among workers, then the swarming begins. The workers start building swarm cells for new virgin queens. When the queen is ready, she leaves the hive and is followed by about 70% of the workers in a massive cloud of flying bees, the swarm (Rangel et al., 2012). The swarm will cluster on a nearby tree and the remain there until they find a suitable new home.

The virgin queens developing in the old hive are daughters of the queen that swarmed and are attended by the remnant workers that did not leave with the swarm. After few days, the new virgin queens begin to emerge. Typically, the first queen to emerge will  kill the rest of virgin queens to assume the role as the new queen for the colony. She will then go on a mating flight to find drones to mate with to begin laying eggs and rebuilding the workforce in the colony.

Clemson Cooperative Extension. (2021). Frequently asked questions about honey bee swarms. <https://hgic.clemson.edu/factsheet/frequently-asked-questions-about-honey-bee-swarms/, accesed on 11.11.2022).
Rangel, J., & Seeley, T. D. (2012). Colony fissioning in honey bees: size and significance of the swarm fraction. Insectes sociaux, 59(4), 453-462.


In the simulation, we first need to initiate a simulation with 10 base virgin queens. We next create a colony (`Colony` class) and an apiary (`MultiColony` class).

```{r create founder genoms and a Colony and Multicolony objects }
founderGenomes <- quickHaplo(nInd = 30, nChr = 1, segSites = 100)
SP <- SimParamBee$new(founderGenomes)
basePop <- createVirginQueens(founderGenomes)
drones <- createDrones(basePop[1], n = 1000)
fatherGroups <- pullDroneGroupsFromDCA(drones, n = 30, nDrones = 10)

# Create Colony and MultiColony class
colony <- createColony(x = basePop[2])
colony <- cross(colony, drones = fatherGroups[[1]])
colony <- buildUp(colony, nWorkers = 100, nDrones = 20)

apiary <- createMultiColony(basePop[3:8], n = 6)
apiary <- cross(apiary, drones = fatherGroups[2:7])
apiary <- buildUp(apiary, nWorkers = 100, nDrones = 20, exact = TRUE)
```

We developed a function `swarm()` that simulates swarming. The function takes a `Colony` class object and a percentage `p` of workers that leave with the swarm. The function returns is a list with two `Colony` clas objects, `swarm` and `remnant`. The `swarm` contains the old queen and `p` percentage of workers that left the hive. The `remnant` contains the rest of workers `(1-p)`, all the drones, and virgin queens that are daughters of the old queen that swarmed. 

Let's swarm our `colony`. Since the output of a function is a list with colonies, we must save it in a new (temporary) object.

```{r swarm colony }
tmp <- swarm(colony, p = 0.4)
tmp
```

The function returns two colonies: `remnant` and `swarm`. The `remnant` colony does not have queen but has one virgin queen that must be mated. It also  has 60 workers since we set `p` argument to 0.4 meaning that 40% of workers will leave with the swarm. All the drones remained in the `remnant` and `Has swarmed` has turned to `TRUE`.

The `swarm` contains the old queen, no virgin queens, and 40 workers, since we set the proportion `p` to 0.4. Same as in the `remnant`, the `Has swarmed` has turned to `TRUE` in the `swarm`.

The `swarm` stays genetically identical to the "old" colony, although downsized. Hence, we assign the `swarm` back to the original colony. The `remnant` has a new queen is hence genetically different from the original colony. Thus, we considered it to be a new colony.

```{r saving output of the function}
colony <- tmp$swarm
colony1 <- tmp$remnant
```

After swarming, the colony would usually build-up back to the full size and the virgin queens would mate.

Instead of setting the `p` everytime we call the `swarm()` function, we can save the `swarmP` argument in the `SimParamBee` object. The `swarm()` function will then use this percentage if `p` is not set.

```{r inspect the p swarm parameter in SP object}
SP$swarmP
```

The default value is 0.5, but we can set any value we want.

```{r set our own swarmP in SP object}
SP$swarmP <- 0.35
SP$swarmP
```

We swarm a `MultiColony` object is the same was we swarm a single `Colony` - with the `swarm()` function. The `swarm() `function is here applied to each colony in the `MultiColony()` object with the same parameters. The function now returns two `MultiColony()` objects - one containing the `swarms` and the other containing the `remnants`.

```{r swarming MultyColony}
tmp <- swarm(apiary)
tmp
```

We see that we get six swarms and six remnants from the apiary with six colonies. We can inspect individuals colonies to ensure they swarmed according to the parameters. Let's inspect the `swarm` and `remnant` of the third colony.

```{r inspect the output (swarm)}
tmp$swarm[[3]]
```

```{r inspect the output (remnant)}
tmp$remnant[[3]]
```

We see that the the third colony was swarmed with `p` of 35% as specified in the `SimParamBee`, hence the swarm left with 35 workers and the old queen and the remnant stayed with a new virgin queen and 65 workers.

Above, all the colonies in a `MultiColony` are swarmed with the same percentage. However, we can also specify a different `p` for each colony.

```{r swarmed MultyColony object with different swarm p}
tmp <- swarm(apiary, p = c(0.3, 0.4, 0.5, 0.6, 0.7, 0.8))
```

If we now inspect the first and the second swarm, we see that each colony has a different percentage of workers that stayed and left
```{r inspect swarmed MultyColony object with different swarm p}
tmp$swarm[[1]]
tmp$swarm[[2]]
```
If you want to track the genetics, you would assign the swarms back to the original apiary and create a new apiary from the remnant. However, if you want to track the position, the remnant actually stay in the same position and would hence be assigned back to the same apiary, while the swarm would be assigned to a new apiary or be lost.

You can also use a non-fixed `p` parameter by using the function `swarmPUnif` that samples the `p` from a uniform distribution between values 0.4 and 0.6 irrespective of colony strength.

```{r assign swarm back to apiary and build up the apiary and colony}
apiary <- tmp$swarm
apiary <- buildUp(apiary)
colony <- buildUp(colony)
```

# Splitting
Colony splitting is a common beekeeping technique for managing swarming. A percentage of workers, brood and food stores are removed and combined in a new colony. Old queen normally stays with splited colony (remnant in our function). Function works on `Colony` and `Multicolony` objects. Argument `p` is proportion of workers that will go to new colony. The output of the function is a list of two `Colony` or `MultiColony` objects.

```{r split a colony}
tmp <- split(colony, p = 0.3)
colony <- tmp$remnant
colony
```

We see that in remnant is old queen (ID 2), and 70% of workers since we set `p` argument to 0.3 meaning that 30% is removed to split. `Has split` is turned to `TRUE`.

```{r inspect the split}
tmp$split
```

In split we have 30% of workers and a virgin queen. `Has split` is turned to `TRUE`. `p` argument can be used as it is demonstrated in `swarm` function. We can specify a vector with different values when spliting `MultiColony` object. `p` argument can be also changed in `SP` object, so we do not specify it each time we call the function. We call `SimParamBee$splitP` to change the `p` argument in the `SP` object. 
```{r split the MultiColony object and inspect the first colony in remnant}
tmp <- split(apiary, p = 0.3)
tmp$remnant[[1]]
```
Now we see that in remnant we have mated queen and 70 workers...
```{r inspect the split of the first colony}
tmp$split[[1]]
```
and a virgin queen with 30 workers in split.
We can use the vector of `p`, different for each colony same as in `swarm()` function.
 
```{r build up the colony and apaiary for further examples}
apiary <- tmp$remnant
apiary <- buildUp(apiary)
colony <- buildUp(colony)
```

# Supersedure

Supersedure is a replacement of a queen by her daughter without interference of beekeeper. Supersedure is a natural way of requeening without swarming. There are many reasons for supersedure: poor physical condition of a queen, old age, diseases, depleted spermatheca, poorly bread queen, reduced pheromone output and many others.

Hamdan, K. (2010). Natural Supersedure of Queens in Honey Bee Colonies. Bee World, 87(3), 52-54.

```{r create colony and multycolony}
colony <- createColony(x = basePop[9])
colony <- cross(colony, drones = fatherGroups[[9]])
colony <- buildUp(colony, nWorkers = 100, nDrones = 20)

apiary <- createMultiColony(basePop[10:15], n = 6)
apiary <- cross(apiary, drones = fatherGroups[10:15])
apiary <- buildUp(apiary, nWorkers = 100, nDrones = 20, exact = TRUE)
```

```{r inspect the colony}
colony
```

```{r supersede Colony object}
colony <- supersede(colony)
colony
```

Now, the old queen is removed and virgin queen ready to mate. We also see that `Has superseded` event is set to `TRUE`. The function `supersede()` works both on `Colony` and `MultiColony` classes. We can select percentage of colonies from 

```{r supersede Multicolony object}
apiary <- supersede(apiary)
apiary[[2]]
```

We do the supersedure of `MultiColony` in a same way as it is described as described above in `Colony` . We don't have to assign the output of the function to a special object since the output is single `Colony` or `MultiColony` object and not a list of two.



# Collapse

Collapse of the colony is a term that describes death of a colony. All individuals within a colony die when collapse occurs. There are many
reasons for collapse of a honey bee colony like: diseases, starvation, queen problems, environmental and in-hive contamination with pesticides.
Colony losses can be very high, up to 60% of colonies per year. High colony losses can significantly influence genetic structure of a population.

```{r create the colony and Mlticolony }
colony <- createColony(x = basePop[20])
colony <- cross(colony, drones = fatherGroups[[20]])
colony <- buildUp(colony, nWorkers = 100, nDrones = 20)

apiary <- createMultiColony(basePop[21:26], n = 6)
apiary <- cross(apiary, drones = fatherGroups[21:26])
apiary <- buildUp(apiary, nWorkers = 100, nDrones = 20, exact = TRUE)
```

```{r collapse the colony}
colony <- collapse(colony)
colony
```

The output of the function is `Colony` with the `Has collapsed` set to `TRUE`. 

```{r collapse the MultiColony}
apiary <- collapse(apiary)
apiary[[3]]
```

The function works also `MultiColony` objects. Colonies from `MultiColony` are selected by using function `pullColonies()` with arguments `n` specifying exact number of collapsed colonies or with `p` argument specifying the percentage of collapsed colonies. `collapse()` function is used when you want to keep collapsed colonies for subsequent genetic analysis. In case when you don't need the collapsed colony you can simply select the survivors with `select()` function.




