---
title: "Colony variance calculus"
output: html_document
date: '2022-11-12'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This document develops theory for the variance of genetic and phenotypic values
between colonies with the aim to understand how to fine tune a SIMplyBee simulation
where we need variance components for queen and workers effects at an individual
honeybee level. The challenge here is that in reality we do not observe such
variances, but only on a colony level.

For example, in the vignette we assumed that phenotypic variance on the individual
level is about 1/4 due to genetic variation and about 3/4 due to environmental
variation, but looking at colony genetic and phenotypic variances the ratio
was aobut 1/2!

TODO: What do we do about this? It feels like we need a full paper on all of this;) :(

To make use of the material in this document you should run the quantitative
genetics vignette - the first example. Then you can continue here!

Having multiple colonies and their values, we can now return to the "about" point
mentioned at the start of this section, when we defined quantitative genetic
parameters. Recall that our starting quantitative genetic parameters for the
queen and workers effects were:

```{r quan_gen_param_revision}
# Trait means
mean

# Trait genetic variation - on a per honeybee level
varA
corA
(covA <- corA * outer(X = sqrt(varA), Y = sqrt(varA), FUN = "*"))

# Trait environmental variation - on a per honeybee level
varE
corE
(covE <- corE * outer(X = sqrt(varE), Y = sqrt(varE), FUN = "*"))

# Trait phenotypic variation - on a per honeybee level
(covP <- covA + covE)
(corP <- cov2cor(covP))
(varP <- diag(covP))

# Expected phenotypic variation - on a per colony level
# TODO: there is in fact more stuff happening here - see below! 
nW <- SP$nWorkers # TODO: this is not correct - we have queen-workers pairs and workers-workers (see below!)
covA[1, 1] + nW * covA[2, 2] + 2 * k * covA[1, 2] # variance can not be negative!
covP[1, 1] + nW * covP[2, 2] + 2 * k * covP[1, 2]

# Observed variation - on a per colony level
var(colonyGv)
var(colonyPheno)
var(colonyGv) / var(colonyPheno) 
```

Hmm, we have much higher genetic and phenotypic variances, as well as much higher
ratio between the two than the initial value of ~1/4! Why? Let's see. For genetic
value of a colony $g_c$ we add up the queen's genetic value for the queen effect
$g_{q,q}$ and workers' genetic values for the workers effect $\Sigma_{i=1}^{n_w}(g_{i,w})$:

$g_c = g_{q,q} + \Sigma_{i=1}^{n_w}(g_{i,w}).$

Expectation of colony genetic value is then: 

$E(g_c) = E(g_{q,q} + \Sigma_{i=1}^{n_w}(g_{i,w})$
$E(g_c) = E(g_{q,q}) + E(\Sigma_{i=1}^{n_w}(g_{i,w}))$
$E(g_c) = \mu_{g_{q}} + n_w E(g_{w,w})$
$E(g_c) = \mu_{g_{q}} + n_w \mu_{g_{w}}$ 

we are assuming here that $n_w$ is a fixed value, but if it is a random variable then we would have 
 
$E(g_c) = \mu_{g_{q}} + E(n_w) E(g_{w,w})$ using https://en.wikipedia.org/wiki/Distribution_of_the_product_of_two_random_variables
$E(g_c) = \mu_{g_{q}} + \lambda_{n_w} \mu_{g_{w}}$ 

which in our case turns the same. So in our case we have:

```{r}
nW <- SP$nWorkers
mean[1] + nW * mean[2]
mean(colonyGv)
mean(colonyPheno)
```

Variance of colony genetic value is then: 

$Var(g_c) = Var(g_{q,q} + \Sigma_{i=1}^_{n_w}(g_{i,w}))$
$Var(g_c) = Var(g_{q,q}) + Var(\Sigma_{i=1}^_{n_w}(g_{i,w})) + 2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w}))$
$Var(g_c) = \sigma^2_{g_{q}} + Var(\Sigma_{i=1}^{n_w}(g_{i,w})) + 2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w}))$

So, we have three parts:
  A) the variance of the queen effect in queens $\sigma^2_{g_{q}}$,
  B) variance of a sum of workers effect in workers $Var(\Sigma_{i=1}^{n_w}(g_{i,w}))$, and
  C) covariance between the two $2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w}))$.

# A) Variance of the queen effect in queens $\sigma^2_{g_{q}}$ - in our case this is:

```{r check_queen_variances}
varA[1]
var(getQueenGv(apiary, collapse = TRUE)[, "queenTrait"])
# ... note that R's var() divides by n-1, which matters with small n
sum((getQueenGv(apiary, collapse = TRUE)[, "queenTrait"] - mean[1])^2) / nColonies(apiary)
```

Anyway, quite close!

# B) Variance of a sum of workers effect in workers $Var(\Sigma_{i=1}^{n_w}(g_{i,w}))$

Given that this is a sum, we need to look at:
  B1) variance of each of worker genetic values $Var(g_{i,w})$ and
  B2) covariance between each pair of the values $Cov(g_{i,w}, g_{j,w})$.
  
This means, that family structure will start to matter because it induces
covariance beteen family members (workers). The simplest case is when all workers
come from the same father. Then we have:

$g_{1,w} = \frac{1}{2}g_{q,w} + g_{f,w} + r_{1,w}$
$g_{2,w} = \frac{1}{2}g_{q,w} + g_{f,w} + r_{2,w}$
...
$g_{n_w,w} = \frac{1}{2}g_{q,w} + g_{f,w} + r_{n_w,w}$

## B1) variance of each of worker genetic values $Var(g_{i,w})$

As stated initially, we have $Var(g_{i,w}) = \sigma^2_{g_{w}}$ and we
have $n_w$ such terms, $n_w \sigma^2_{g_{w}}$. In our case this would be:

```{r check_workers_variances}
nW * varA[2]
g <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = "workersTrait")
var(g)
# ... note that R's var() divides by n-1, which matters with small n
sum((g - nW * mean[2])^2) / nColonies(apiary)
```

So, a huge discrepancy between the variance of a sum of genetic values for the
workers effect `nW * varA[2]` (assuming independence) and a realised
sum of genetic values for the workers effect. As we have seen above, this is
due to family structure and associated B2) covariance between each pair of the
values $Cov(g_{i,w}, g_{j,w})$.

## B2) covariance between each pair of the values $Cov(g_{i,w}, g_{j,w})$

Clearly, these covariances matter a lot - there is lots of worker pairs in a
colony! How do these covariances look like?

$Cov(g_{1,w}, g_{2,w}) = Cov(\frac{1}{2}g_{q,w} + g_{f,w}, \frac{1}{2}g_{q,w} + g_{f,w})$
$Cov(g_{1,w}, g_{2,w}) = Var(\frac{1}{2}g_{q,w}) + Var(g_{f,w})$

assuming that mother and father are not related; further genetic variance between
drones is in fact half the genetic variance between queens because they are haploid,
so we get:

$Cov(g_{1,w}, g_{2,w}) = \frac{1}{4}\sigma^2_{g_w} + \frac{1}{4}\sigma^2_{g_w}$
$Cov(g_{1,w}, g_{2,w}) = \frac{1}{2}\sigma^2_{g_w}$.

Every pair of super-sister workers adds $\frac{1}{2}\sigma^2_{g_w}$ to B - note
that pair A-B adds this value, but so does the pair B-A, hence total is
$2\frac{1}{2}\sigma^2_{g_w} = \sigma^2_{g_w}$.
With $n_w$ workers we get $n_w n_w$ pairs (including with itself) or
$n_w n_w - n_w$ pairs between different workers. The total covariance
contribution is then $(\frac{n_w n_w}{2} - \frac{n_w}{2})\sigma^2_{g_w}$
(or equivalently $(n_w n_w - n_w)\frac{1}{2}\sigma^2_{g_w}$). In our case
this would be:

```{r check_workers_variances2} 
(varSumWorkers <- nW * varA[2]) # B1
(covSumWorkers <- (nW * nW - nW)/2 * varA[2]) # B2
varSumWorkers + covSumWorkers # B

g <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = "workersTrait")
var(g) # B
# ... note that R's var() divides by n-1, which matters with small n
sum((g - nW * mean[2])^2) / nColonies(apiary) # B
```

If we would only have super-sisters we would have ~1 for B1, ~49.5 for B2, and
~50.5 for B, but with a mix of sisters and super-sisters we would have a lower
value, which is what we see above, ~26.8 for B.

So, now we need to work out B2 with a mix of sisters and super-sisters :( 
Sister workers are related only due to having the same mother, assuming that
drones the queen mated with are unrelated:

$g_{1,w} = \frac{1}{2}g_{q,w} + g_{f1,w} + r_{1,w}$
$g_{2,w} = \frac{1}{2}g_{q,w} + g_{f2,w} + r_{2,w}$

$Cov(g_{1,w}, g_{2,w}) = Cov(\frac{1}{2}g_{q,w} + g_{f1,w}, \frac{1}{2}g_{q,w} + g_{f2,w})$
$Cov(g_{1,w}, g_{2,w}) = Var(\frac{1}{2}g_{q,w})$
$Cov(g_{1,w}, g_{2,w}) = \frac{1}{4}Var(g_{q,w})$

Hence every pair of sister workers adds $\frac{1}{4}\sigma^2_{g_w}$ to B (note,
this is $2\frac{1}{4}\sigma^2_{g_w}$ for A-B and B-A pairs!).
With $n_w$ workers and $n_f$ fathers we have $\frac{n_w}{n_f}$ workers per father.
Further, we have $n_f$ groups of super-sisters and $n_f n_f - n_f$ pairs of 
sister groups. Assuming that sister groups are the same size, we have
$n_f \frac{n_w}{n_f} \frac{n_w}{n_f} = n_w \frac{n_w}{n_f} = \frac{n^2_w}{n_f}$ pairs of
super-sisters and $(n_f n_f - n_f) \frac{n_w}{n_f} \frac{n_w}{n_f}$ - this is 
$n_f (n_f - 1) \frac{n_w}{n_f} \frac{n_w}{n_f} = (n_f - 1) n_w \frac{n_w}{n_f} = \frac{ (n_f - 1) n^2_w}{n_f}$. Is this correct?

```{r pairs_of_workers}
(nF <- nFathers(colony))
nW * nW
(a <- nW * nW / nF)
(b <- (nF - 1) * nW * nW / nF)
a+b 
```

Looks like!

So, super-sister pairs add $\frac{n^2_w}{n_f} \frac{1}{2}\sigma^2_{g_w}$ and
sister pairs add $\frac{ (n_f - 1) n^2_w}{n_f} \frac{1}{4}\sigma^2_{g_w}$. Let's
test this:

```{r check_workers_variances3}
(varSumWorkers <- nW * varA[2]) # B1
(covSumWorkersSuperSisters <- (nW * nW / nF) * 0.50 * varA[2]) # B2 in super-sisters
(covSumWorkersSisters <- ((nF - 1) * nW * nW / nF) * 0.25 * varA[2]) # B2 in super-sisters
varSumWorkers + covSumWorkersSuperSisters + covSumWorkersSisters # B

g <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = "workersTrait")
var(g) # B
# ... note that R's var() divides by n-1, which matters with small n
sum((g - SP$nWorkers * mean[2])^2) / nColonies(apiary) # B
```

Cool - very close!!!

TODO: What if $n_w$ is a random variable? Following  https://en.wikipedia.org/wiki/Distribution_of_the_product_of_two_random_variables
we would use:

$Var(XY) = (Var(X) + E(X)^2) (Var(Y) + E(Y)^2) - E(X)^2 E(Y)^2$ 

but this is for two variables, while I have $n_w$ genetic values, so this reference
result is not that useful in our case! Well, if $n_w$ varies between colonies,
this must boost variance of colony-level genetic values significantly - because
kind of start calculating variance between apples (small colonies) and oranges
(large colonies).

# C) Covariance between the two $2Cov(g_{q,q}, \Sigma_{i-1}^{n_w}(g_{i,w}))$.

Let's repeat:

$g_{1,w} = \frac{1}{2}g_{q,w} + g_{f1,w} + r_{1,w}$
$g_{2,w} = \frac{1}{2}g_{q,w} + g_{f2,w} + r_{2,w}$
...

The essential bit here is:

$Cov(g_{q,q}, g_{i,w}) = Cov(g_{q,q}, \frac{1}{2} g_{q,w})$

assuming that the queen and fathers are unrelated. Then:

$Cov(g_{q,q}, g_{i,w}) = Cov(g_{q}, \frac{1}{2} g_{w})$
$Cov(g_{q,q}, g_{i,w}) = \frac{1}{2} Cov(g_{q}, g_{w})$
$Cov(g_{q,q}, g_{i,w}) = \frac{1}{2} \sigma_{g_{q},g_{w}}$

So:

$2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w})) = 2Cov(g_{q,q}, n_w \frac{1}{2} g_{q,w})$
$2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w})) = 2 n_w \frac{1}{2}Cov(g_{q,q}, g_{q,w})$
$2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w})) = n_w \sigma_{g_{q},g_{w}}$

```{r check_workers_variances4}
(varQueen <- varA[1]) # A

(varSumWorkers <- nW * varA[2]) # B1
(covSumWorkersSuperSisters <- (nW * nW / nF) * 0.50 * varA[2]) # B2 in super-sisters
(covSumWorkersSisters <- ((nF - 1) * nW * nW / nF) * 0.25 * varA[2]) # B2 in sisters
(varSumWorkers <- varSumWorkers + covSumWorkersSuperSisters + covSumWorkersSisters) # B

(covQueenSumWorkers <- nW * covA[1, 2]) # C

varQueen + varSumWorkers + covQueenSumWorkers

g <- colonyGv
var(g)
sum((g - 20)^2) / nColonies(apiary)


(varQueen <- varP[1]) # A

(varSumWorkers <- nW * varP[2]) # B1
(covSumWorkersSuperSisters <- (nW * nW / nF) * 0.50 * varA[2]) # B2 in super-sisters - only genetic cov here, I think TODO!?
(covSumWorkersSisters <- ((nF - 1) * nW * nW / nF) * 0.25 * varA[2]) # B2 in sisters - only genetic cov here, I think TODO!?
(varSumWorkers <- varSumWorkers + covSumWorkersSuperSisters + covSumWorkersSisters) # B

(covQueenSumWorkers <- nW * covP[1, 2]) # C - pheno or genetic only here - TODO!?

varQueen + varSumWorkers + covQueenSumWorkers

p <- colonyPheno
var(p)
sum((p - 20)^2) / nColonies(apiary)
```

Bingo for genetic variance and very close for phenotypic too, but we need to check some part above!

# What now?

Alright, what can we now do with this? We now have a system of expected genetic
and phenotypic variances for colony values as a function of number of workers,
fathers, and genetic and phenotypic covariances between the queen and worker
effects. So, it should be possible, in principle, to solve for what kind of
covariance values for the queen and workers effects should we use, to get the
desired colony genetic and phenotypic variances!

TODO: develop such estimating equations!

So, we have genetic part:

$A_g = \sigma^2_{g_q}$
$B1_g = n_w \sigma^2_{g_w}$
$B2_g = \frac{n^2_w}{n_f} \frac{1}{2} \sigma^2_{g_w}$
$B3_g = \frac{(n_f - 1) n^2_w}{n_f} \frac{1}{4} \sigma^2_{g_w}$
$C_g = n_w \sigma^2_{{g_q},{g_w}}$

here are 3 unknowns: $\sigma^2_{g_q}$, $\sigma^2_{g_w}$, and $\sigma^2_{{g_q},{g_w}}$

we can add phenotypic part:

$A_p = \sigma^2_{p_q}$
$B1_p = n_w \sigma^2_{p_w}$
$B2_p = \frac{n^2_w}{n_f} \frac{1}{2} \sigma^2_{g_w}$ TODO: check if its only genetic
$B3_p = \frac{(n_f - 1) n^2_w}{n_f} \frac{1}{4} \sigma^2_{g_w}$ TODO: check if its only genetic
$C_p = n_w \sigma^2_{{p_q},{p_w}}$

here are additional 3 unknowns: $\sigma^2_{p_q}$, $\sigma^2_{p_w}$, and $\sigma^2_{{p_q},{p_w}}$

while we could provide as guidance 2, maybe 3, inputs: $\sigma^2_{g_c}$, $\sigma^2_{p_c}$, and $\frac{\sigma^2_{g_c}}{\sigma^2_{p_c}}$.

TODO: are there any additional inputs that we could get from literature? How do
we connect all this with variance estimates for the queen and the workers effect
in the honeybee genetics and breeding literature? I think that those estimates are
essentially this: $\sigma^2_{g_c} = \sigma^2_{g_{c,q}} + \sigma^2_{g_{c,w}} + 2\sigma_{g_{c,q},g_{c,w}}$ where $\sigma^2_{g_{c,q}}$ is the queen part of $\sigma^2_{g_c}$,
but not necessarily the same as $\sigma^2_{g_w}$, or is it? Maybe that is the same,
and $\sigma^2_{g_{c,w}}$ is a counterpart to $n_w \sigma^2_{g_w} + \frac{n^2_w}{n_f} \frac{1}{2} \sigma^2_{g_w}, \frac{(n_f - 1) n^2_w}{n_f} \frac{1}{4} \sigma^2_{g_w}$, while
$\sigma_{g_{c,q},g_{c,w}}$ is a counterpart to $n_w \sigma^2_{{g_q},{g_w}}$. If
this is so, then it fells that we should have 3 more inputs for genetic part -
so 4 inputs to estimate 3 unknowns of which some are very simple transformations!!!

Additional input could be $\sigma^2_{p_c}$. For sure we don't get estimates of 
$\sigma^2_{p_c} = \sigma^2_{p_{c,q}} + \sigma^2_{p_{c,w}} + 2\sigma_{p_{c,q},p_{c,w}}$
but, having sorted the genetic part and if $B2_p$ and $B3_p$ depend only on
genetic covariance, then maybe, we have some maneuvering space to estimate also
$\sigma^2_{p_q}$, $\sigma^2_{p_w}$, and $\sigma^2_{{p_q},{p_w}}$?!

Furthermore, we have some constraints, so this is not just any system of equations.
For example, we know that variances must be positive. Futhermore, we might know
that come covariances are either positive or negative.

TODO: Develop and polish all this further and create a Shiny application where
we can change these individual-level parameters and study how that will change
colony-level parameters and then how we could go the other way around based on
literature and solving the above systems of equations. We could host such a
Shiny app on SIMplyBee.info!? Can we?
