---
  title: "Variance calculations between individual and colony level values"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{VarianceCalculations}
%\VignetteEngine{knitr::rmarkdown}
%\VignetteEncoding{UTF-8}
editor_options:
  markdown:
  wrap: 80
canonical: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE
)
```

```{r cor2cov, include = FALSE}
cor2cov <- function(corr, var) {
  # Ensure inputs are matrices/vectors
  corr <- as.matrix(corr)
  var  <- as.numeric(var)

  # Standard deviations
  sd <- sqrt(var)

  # Covariance matrix
  cov <- corr * (sd %o% sd)

  return(cov)
}
```

Let's again run the example from the quantitative genetic vignette.

```{r FirstExampleFromQuanGenVignette}
# Run the example from the quantitative genetic vignette
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
nQtlPerChr <- 100
mean <- c(10, 10 / SP$nWorkers)
varA <- c(1, 1 / SP$nWorkers)
corA <- matrix(data = c( 1.0, -0.5,
                         -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
             name = c("queenTrait", "workersTrait"))
varE <- c(3, 3 / SP$nWorkers)
corE <- matrix(data = c(1.0, 0.3,
                        0.3, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)
head(basePop@gv)
head(basePop@pheno)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)
colonyGv <- calcColonyGv(apiary)
colonyPheno <- calcColonyPheno(apiary)

```

How the expected and realised variances look like
```{r}
k <- 1
print("Expected genetic and phenotypic variance")
covA[1, 1] + nW * covA[2, 2] + 2 * k * covA[1, 2] # variance can not be negative!
covP[1, 1] + nW * covP[2, 2] + 2 * k * covP[1, 2]

print("Realised genetic and phenotypic variance")
var(colonyGv)
var(colonyPheno)
```

Now, in theory, how we go from the individual to colony level variances?
The queen trait variance remains the same, since there is only one queen. The worker
individual variance needs to be adjusted due to covariances between workers.

```{r}
a_ii = 0.40
varA_colony <- c(varA[1], varA[2] * a_ii)
corA = cov2cor(covA)
covA_colony = cor2cov(corA, var = varA_colony)

covA[1, 1] + nW * covA[2, 2] + 2 * k * covA[1, 2]
covA_colony[1, 1] + covA_colony[2, 2] + 2 * k * covA_colony[1, 2]

var(colonyGv)
var(colonyPheno)
```

We need to add the covariance due to different types of workers in the colony
```{r}
# Get the real number of fathers and DPQs
nFathers(apiary)
median(sapply(getFathers(apiary), FUN = function(x) length(unique(x@father))))

# Genetic values
nW = SP$nWorkers #100 #BB assumes mean - so calculate as if this is 1
nF = SP$nFathers
nDPQ = 5

varA_q <- 1
varA_w = 1 / SP$nWorkers
corA_qw = corA[1,2]

# Covariance due to super-sisters
covSumWorkersSuperSisters <- (nW * nW / nF) * 0.75 * varA_w
# Covariance due to full-sisters
covSumWorkersFullSisters <- ((nW * nW / nDPQ) -  (nW * nW / nF)) * 0.50 * varA_w
# Covariance due to half-sisters
covSumWorkersHalfSisters <- ((nF - 1) * nW * nW / nF) * 0.25 * varA_w




mapIndToColonyVar <- function(varA_q, varA_w, corA_qw, nW, nF, nDPQ, workersFUN = "sum") {
  if (workersFUN == "mean") {
    nW = 1
  }
  
  # Determine how many pairs of each you have
  n_SS = nW * nW / nF
  n_FS = (nW * nW / nDPQ) -  (nW * nW / nF)
  n_HS = (nW * nW / nDPQ) * (nDPQ - 1)
  A <- varA_q
  B1 <- nW * varA_w
  B2_ss <- n_SS * 0.75 * varA_w
  B2_fs <- n_FS * 0.50 * varA_w
  B2_hs <- n_HS * 0.25 * varA_w
  
  covA_qw = corA_qw * sqrt(varA_q) * sqrt(varA_w)
  C <- nW * covA_qw
  
  if (workersFUN == "sum") {
    varC <- A + B1 + B2_ss + B2_fs + B2_hs + C
  } else if (workersFUN == "mean") {
    stop("TODO: derive this equation properly")
    #varC <- varA_q + varA_w + (B2_ss + B2_fs + B2_hs) + covA_qw # This is equal as varA_w * average relatedness within a colony
  }
  return(varC)
}

#This is the same as above, just simplified
mapIndToColonyVar <- function(varA_q, varA_w, corA_qw, nW, nF, nDPQ, workersFUN = "sum") {

  # Determina how many pairs of each you have
  p_SS = 1 / nF
  p_FS = (1 / nDPQ) -  (1 / nF)
  p_HS = (1 / nDPQ) * (nDPQ - 1)
  A <- varA_q
  B1 <- nW * varA_w
  B2_ss <- p_SS * 0.75
  B2_fs <- p_FS * 0.50
  B2_hs <- p_HS * 0.25
  
  covA_qw = corA_qw * sqrt(varA_q) * sqrt(varA_w)
  C <- nW * covA_qw
  
  if (workersFUN == "sum") {
    varC <- A + B1 + (B2_ss + B2_fs + B2_hs)*nW^2*varA_w + C
  } else if (workersFUN == "mean") {
    varC <- A + varA_w + (B2_ss + B2_fs + B2_hs)*varA_w + covA_qw# This is equal as varA_w * average relatedness within a colony
  }
  return(varC)
}

```

```{r}
# When the function to summarise worker effects is SUM
mapIndToColonyVar(varA_q = varA_q, 
                  varA_w= varA_w,
                  corA_qw = corA_qw,
                  nW = nW, nF = nF,
                  nDPQ = nDPQ,
                  workersFUN = "sum")
colonyGv <- calcColonyGv(apiary, FUN = mapCasteToColonyGv, workersFUN = colSums)
var(colonyGv)
```

```{r}
# When the function to summarise worker effects is MEAN
mapIndToColonyVar(varA_q = varA_q, 
                  varA_w = varA_w * SP$nWorkers,
                  corA_qw = corA_qw,
                  nW = nW, nF = nF,
                  nDPQ = nDPQ,
                  workersFUN = "mean")
colonyGv <- calcColonyGv(apiary, FUN = mapCasteToColonyGv, workersFUN = colMeans)
var(colonyGv)
```
```

```{r}
# Try to frame something for the other way around
# At best, we would have varA_q (colony), varA_w (colony),

```


