---
title: "Variance calculations between individual and colony level values - functions"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VarianceCalculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 80
    canonical: true
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE
)
```

```{r cor2cov, include = FALSE}
library(package = "SIMplyBee")
cor2cov <- function(corr, var) {
  # Ensure inputs are matrices/vectors
  corr <- as.matrix(corr)
  var  <- as.numeric(var)

  # Standard deviations
  sd <- sqrt(var)

  # Covariance matrix
  cov <- corr * (sd %o% sd)

  return(cov)
}
```

Let's again run the example from the quantitative genetic vignette.

```{r FirstExampleFromQuanGenVignette}
# Run the example from the quantitative genetic vignette
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
nQtlPerChr <- 100
mean <- c(10, 10 / SP$nWorkers)
varA <- c(1, 1 / SP$nWorkers)
corA <- matrix(data = c( 1.0, -0.5,
                         -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
             name = c("queenTrait", "workersTrait"))
varE <- c(3, 3 / SP$nWorkers)
corE <- matrix(data = c(1.0, 0.3,
                        0.3, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)
head(basePop@gv)
head(basePop@pheno)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)
colonyGv <- calcColonyGv(apiary)
colonyPheno <- calcColonyPheno(apiary)

```

How the expected and realised variances look like
```{r ExpectedAndRealisedVariances}
corA
(covA <- corA * outer(X = sqrt(varA), Y = sqrt(varA), FUN = "*"))

# Trait environmental variation - on a per honeybee level
(covE <- corE * outer(X = sqrt(varE), Y = sqrt(varE), FUN = "*"))

# Trait phenotypic variation - on a per honeybee level
(covP <- covA + covE)
(corP <- cov2cor(covP))
(varP <- diag(covP))
nW <- SP$nWorkers


k <- 1
print("Expected genetic and phenotypic variance")
covA[1, 1] + nW * covA[2, 2] + 2 * k * covA[1, 2] # variance can not be negative!
covP[1, 1] + nW * covP[2, 2] + 2 * k * covP[1, 2]

print("Realised genetic and phenotypic variance")
var(colonyGv)
var(colonyPheno)
```

Now, in theory, how we go from the individual to colony level variances?
The queen trait variance remains the same, since there is only one queen. The worker
individual variance needs to be adjusted due to covariances between workers.

```{r AdjustingWorkerVariance}
a_ii = 0.40
varA_colony <- c(varA[1], varA[2] * a_ii)
corA = cov2cor(covA)
covA_colony = cor2cov(corA, var = varA_colony)

covA[1, 1] + nW * covA[2, 2] + 2 * k * covA[1, 2]
covA_colony[1, 1] + covA_colony[2, 2] + 2 * k * covA_colony[1, 2]

var(colonyGv)
var(colonyPheno)
```

We need to add the covariance due to different types of workers in the colony
```{r CovariancesBetweenWorkers}
mapIndToColonyVar <- function(varA_q, varA_w, corA_qw, 
                              varE_q, varE_w, corE_qw,
                              nW, nF, nDPQ, workersFUN = "sum") {
  
  # First handle the genetic part
  # Determine how many pairs of each you have
  nW = nW
  n_SS = (nW * nW / nF) - nW
  n_FS = (nW * nW / nDPQ) -  (nW * nW / nF)
  n_HS = (nW * nW / nDPQ) * (nDPQ - 1)

  varA_q <- varA_q
  if (workersFUN == "mean") {
    B1 = 1 / nW * varA_w
  } else if (workersFUN == "sum") {
    B1 = nW * varA_w
  }

  B2_ss <- n_SS * 0.75 * varA_w
  B2_fs <- n_FS * 0.50 * varA_w
  B2_hs <- n_HS * 0.25 * varA_w

  if (workersFUN == "mean") {
    varA_wbar = B1 + 1/nW^2 * (B2_ss + B2_fs + B2_hs)
  } else if (workersFUN == "sum") {
    varA_wbar = B1 + B2_ss + B2_fs + B2_hs
  }

  covA_qw = corA_qw * sqrt(varA_q) * sqrt(varA_w)

  if (workersFUN == "mean") {
    covA_qwbar <- covA_qw
  } else if (workersFUN == "sum") {
    covA_qwbar <- nW * covA_qw
  }
  corA_qwbar <- covA_qwbar / (sqrt(varA_q) * sqrt(varA_wbar))

  varA_c <- varA_q + varA_wbar + 2*covA_qwbar

  # Next handle the environmental part
  if (workersFUN == "mean") {
    varE_wbar = 1 / nW * varE_w
  } else if (workersFUN == "sum") {
    varE_wbar = nW * varE_w
  }

  covE_qw = corE_qw * sqrt(varE_q) * sqrt(varE_w)

  if (workersFUN == "mean") {
    covE_qwbar <- covE_qw
  } else if (workersFUN == "sum") {
    covE_qwbar <- nW * covE_qw
  }
  corE_qwbar <- covE_qwbar / (sqrt(varE_q) * sqrt(varE_wbar))

  varE_c <- varE_q + varE_wbar + 2*covE_qwbar

  return(list(varA_q = varA_q, varA_wbar = varA_wbar, covA_qwbar = covA_qwbar, corA_qwbar = corA_qwbar, varA_c = varA_c,
              varE_q = varE_q, varE_wbar = varE_wbar, covE_qwbar = covE_qwbar, corE_qwbar = corE_qwbar, varE_c = varE_c))
}
```

```{r testingMapIndToColonyVar}
print("Workers aggregate is a mean")
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
nQtlPerChr <- 100
mean <- c(0, 0)
varA <- c(1, 1)
corA <- matrix(data = c( 1.0, -0.5,
                        -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
            name = c("queenTrait", "workersTrait"))
varE <- c(3, 3)
corE <- matrix(data = c(1.0, 0.3,
                        0.3, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)
head(basePop@gv)
head(basePop@pheno)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)
colonyGv <- calcColonyGv(apiary)

# Get the real number of fathers and DPQs
nW = round(mean(nWorkers(apiary)), 0)
nF = round(mean(nFathers(apiary)), 0)
nDPQ <- round(mean(sapply(getFathers(apiary), function(x) length(unique(x@mother)))), 0)


# When the function to summarise worker effects is SUM
calcVar <- mapIndToColonyVar(varA_q = varA[1], 
                  varA_w= varA[2],
                  corA_qw = corA[1,2],
                  varE_q = varE[1], 
                  varE_w = varE[2], 
                  corE_qw = corE[1,2],
                  nW = nW, nF = nF,
                  nDPQ = nDPQ,
                  workersFUN = "mean")

real_gv_q <- calcColonyGv(apiary, FUN = mapCasteToColonyGv, queenTrait = 1, workersTrait = NULL)
real_gv_wbar <- calcColonyGv(apiary, FUN = mapCasteToColonyGv, queenTrait = NULL, workersTrait = 2, workersFUN = colMeans)
real_covA_qwbar <- cov(real_gv_q, real_gv_wbar)
real_corA_qwbar <- cor(real_gv_q, real_gv_wbar)
real_varA_q <- var(real_gv_q)
real_varA_wbar <- var(real_gv_wbar)
real_varA_c <- var(calcColonyGv(apiary, FUN = mapCasteToColonyGv, workersFUN = colMeans))
real_varP_c <- var(calcColonyPheno(apiary, FUN = mapCasteToColonyPheno, workersFUN = colMeans))

variances0 <- rbind(variances0, data.frame(rep = rep, 
                                            calc_varA_q = calcVar$varA_q, calc_varA_wbar = calcVar$varA_wbar, calc_varA_c = calcVar$varA_c, calc_varE_c = calcVar$varE_c,
                                            real_varA_q = real_varA_q[1,1], real_varA_wbar = real_varA_wbar[1,1], real_varA_c = real_varA_c[1,1],
                                            real_covA_qwbar = real_covA_qwbar[1,1], real_corA_qwbar = real_corA_qwbar[1,1],
                                            real_varP_c = real_varP_c[1,1],
                                            fun = "mean"))

print("Workers aggregate is a sum")
# Run the example from the quantitative genetic vignette
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
SP$nWorkers <- 100
SP$nFathers <- 15
nQtlPerChr <- 100
mean <- c(0, 0)
varA <- c(1, 1 / SP$nWorkers)
corA <- matrix(data = c( 1.0, -0.5,
                        -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
            name = c("queenTrait", "workersTrait"))
varE <- c(3, 3 / SP$nWorkers)
corE <- matrix(data = c(1.0, 0.3,
                        0.3, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)

basePop <- createVirginQueens(founderGenomes, n = 20)
head(basePop@gv)
head(basePop@pheno)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)
colonyGv <- calcColonyGv(apiary)
colonyPheno <- calcColonyPheno(apiary)  

nW = round(mean(nWorkers(apiary)), 0)
nF = round(mean(nFathers(apiary)), 0)
nDPQ <- round(mean(sapply(getFathers(apiary), function(x) length(unique(x@mother)))), 0)


# When the function to summarise worker effects is SUM
calcVar <- mapIndToColonyVar(varA_q = varA[1], 
                  varA_w= varA[2],
                  corA_qw = corA[1,2],
                  varE_q = varE[1],
                  varE_w = varE[2],
                  corE_qw = corE[1,2],
                  nW = nW, nF = nF,
                  nDPQ = nDPQ,
                  workersFUN = "sum")

real_gv_q <- calcColonyGv(apiary, FUN = mapCasteToColonyGv, queenTrait = 1, workersTrait = NULL)
real_gv_wbar <- calcColonyGv(apiary, FUN = mapCasteToColonyGv, queenTrait = NULL, workersTrait = 2, workersFUN = colSums)
real_covA_qwbar <- cov(real_gv_q, real_gv_wbar)
real_corA_qwbar <- cor(real_gv_q, real_gv_wbar)
real_varA_q <- var(real_gv_q)
real_varA_wbar <- var(real_gv_wbar)
real_varA_c <- var(calcColonyGv(apiary, FUN = mapCasteToColonyGv, workersFUN = colSums))
real_varP_c <- var(calcColonyPheno(apiary, FUN = mapCasteToColonyPheno, workersFUN = colMeans))


variances0 <- rbind(variances0, data.frame(rep = rep, 
                                            calc_varA_q = calcVar$varA_q, calc_varA_wbar = calcVar$varA_wbar, calc_varA_c = calcVar$varA_c, calc_varE_c = calcVar$varE_c,
                                            real_varA_q = real_varA_q[1,1], real_varA_wbar = real_varA_wbar[1,1], real_varA_c = real_varA_c[1,1],
                                            real_covA_qwbar = real_covA_qwbar[1,1], real_corA_qwbar = real_corA_qwbar[1,1],
                                            real_varP_c = real_varP_c[1,1],
                                            fun = "sum"))
```

Now the other way around. Let's say you have a colony-level estimate of variance components - this is what you would get out of a paper.

```{r mapColonyToIndVar}
# First option, we know the varA_q, varA_wbar and cor, and one varE
mapColonyToIndVar <- function(varA_q,
                              varA_wbar,
                              corA_qwbar,
                              varE_q,
                              varE_wbar,
                              corE_qwbar,
                              nW,
                              nF,
                              nDPQ,
                              workersFUN = "sum") {
  
  # First handle the genetic part
  covA_qwbar <- corA_qwbar * sqrt(varA_q) * sqrt(varA_wbar)

  # scaling factor
  if (workersFUN == "sum") {
    covA_qw <- covA_qwbar / nW 
  } else if (workersFUN == "mean") {
    covA_qw <- covA_qwbar
  }

  # pair counts
  n_SS <- (nW * nW / nF) - nW
  n_FS <- (nW * nW / nDPQ) - (nW * nW / nF)
  n_HS <- (nW * nW / nDPQ) * (nDPQ - 1)

  # worker variance coefficient
  if (workersFUN == "sum") {
  K <- nW +
        n_SS * 0.75 +
        n_FS * 0.50 +
        n_HS * 0.25
  } else if (workersFUN == "mean") {
    K <- 1 / nW + ((n_SS * 0.75 + 
                    n_FS * 0.50 + 
                    n_HS * 0.25) / nW^2)
  }

  varA_w <- varA_wbar / K

  corA_qw <- covA_qw / (sqrt(varA_q) * sqrt(varA_w))

  # Next handle the environmental part
  covE_qwbar <- corE_qwbar * sqrt(varE_q) * sqrt(varE_wbar)

  if (workersFUN == "sum") {
    varE_w <- varE_wbar / nW
    covE_qw <- covE_qwbar / nW
  } else if (workersFUN == "mean") {
    varE_w <- varE_wbar * nW
    covE_qw <- covE_qwbar
  }

  corE_qw <- covE_qw / (sqrt(varE_q) * sqrt(varE_w))


  return(list(varA_q = varA_q, varA_wbar = varA_wbar, varA_w = varA_w, 
              covA_qwbar = covA_qwbar, covA_qw = covA_qw, corA_qw = corA_qw,
              varE_q = varE_q, varE_wbar = varE_wbar, varE_w = varE_w,
              covE_qwbar = covE_qwbar, covE_qw = covE_qw, corE_qw = corE_qw))
}
```

```{r testMapColonyToIndFunction}
varA_q <- 5
varA_wbar <- 10
corA_qwbar <- -0.5

varE_q <- 10
varE_wbar <- 20
corE_qwbar <- 0.3

nW <- 100
nF <- 15
nDPQ <- 5

print("Workers aggregate is a mean")
indVarComp <- mapColonyToIndVar(varA_q = varA_q, varA_wbar = varA_wbar, corA_qwbar = corA_qwbar, 
                                varE_q = varE_q, varE_wbar = varE_wbar, corE_qwbar = corE_qwbar, 
                                nW = nW, nF = nF, nDPQ = nDPQ, workersFUN = fun)
calc_var_c <- mapIndToColonyVar(varA_q = indVarComp$varA_q, varA_w = indVarComp$varA_w, corA_qw = indVarComp$corA_qw, 
                                varE_q = indVarComp$varE_q, varE_w = indVarComp$varE_w, corE_qw = indVarComp$corE_qw,
                                nW = nW, nF = nF, nDPQ = nDPQ, workersFUN = fun)


founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
SP$nWorkers = nW
SP$nFathers = nF
nQtlPerChr <- 100
mean <- c(0, 0)
varA <- c(indVarComp$varA_q, indVarComp$varA_w)
corA <- matrix(data = c( 1.0, indVarComp$corA_qw,
                        indVarComp$corA_qw,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
            name = c("queenTrait", "workersTrait"))
varE <- c(indVarComp$varE_q, indVarComp$varE_w)
corE <- matrix(data = c(1.0, indVarComp$corE_qw,
                        indVarComp$corE_qw, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)

drones <- createDrones(x = basePop[1:nDPQ], nInd = 3)
colony <- createColony(x = basePop[nDPQ+1])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:nDPQ], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)

if (fun == "sum") {
  workersFUN <- colSums
} else if (fun == "mean") {
  workersFUN <- colMeans
}

# Get the real number of fathers and DPQs
real_A_wbar <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = 2, workersFUN = workersFUN)
real_varA_wbar <- popVar(real_A_wbar)
real_A_c <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = 1, workersTrait = 2, workersFUN = workersFUN)
real_varA_c <- popVar(real_A_c)
real_P_wbar <- calcColonyPheno(apiary, mapCasteToColonyPheno, queenTrait = NULL, workersTrait = 2, workersFUN = workersFUN)
real_E_wbar <- real_P_wbar - real_A_wbar
real_varE_wbar <- popVar(real_E_wbar)
real_P_c <- calcColonyPheno(apiary, mapCasteToColonyPheno, queenTrait = 1, workersTrait = 2, workersFUN = workersFUN)
real_E_c <- real_P_c - real_A_c
real_varE_c <- popVar(real_E_c)


print("Workers aggregate is a sum")
fun = "sum"
indVarComp <- mapColonyToIndVar(varA_q = varA_q, varA_wbar = varA_wbar, corA_qwbar = corA_qwbar, 
                                varE_q = varE_q, varE_wbar = varE_wbar, corE_qwbar = corE_qwbar, 
                                nW = nW, nF = nF, nDPQ = nDPQ, workersFUN = fun)
calc_var_c <- mapIndToColonyVar(varA_q = indVarComp$varA_q, varA_w = indVarComp$varA_w, corA_qw = indVarComp$corA_qw, 
                                varE_q = indVarComp$varE_q, varE_w = indVarComp$varE_w, corE_qw = indVarComp$corE_qw,
                                nW = nW, nF = nF, nDPQ = nDPQ, workersFUN = fun)

founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
SP$nWorkers = nW
SP$nFathers = nF
nQtlPerChr <- 100
mean <- c(0, 0)
varA <- c(indVarComp$varA_q, indVarComp$varA_w)
corA <- matrix(data = c( 1.0, indVarComp$corA_qw,
                        indVarComp$corA_qw,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
            name = c("queenTrait", "workersTrait"))
varE <- c(indVarComp$varE_q, indVarComp$varE_w)
corE <- matrix(data = c(1.0, indVarComp$corE_qw,
                        indVarComp$corE_qw, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)

drones <- createDrones(x = basePop[1:nDPQ], nInd = 3)
colony <- createColony(x = basePop[nDPQ+1])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:nDPQ], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)

if (fun == "sum") {
  workersFUN <- colSums
} else if (fun == "mean") {
  workersFUN <- colMeans
}

# Get the real number of fathers and DPQs
real_A_wbar <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = 2, workersFUN = workersFUN)
real_varA_wbar <- popVar(real_A_wbar)
real_A_c <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = 1, workersTrait = 2, workersFUN = workersFUN)
real_varA_c <- popVar(real_A_c)
real_P_wbar <- calcColonyPheno(apiary, mapCasteToColonyPheno, queenTrait = NULL, workersTrait = 2, workersFUN = workersFUN)
real_E_wbar <- real_P_wbar - real_A_wbar
real_varE_wbar <- popVar(real_E_wbar)
real_P_c <- calcColonyPheno(apiary, mapCasteToColonyPheno, queenTrait = 1, workersTrait = 2, workersFUN = workersFUN)
real_E_c <- real_P_c - real_A_c
real_varE_c <- popVar(real_E_c)
```
