---
title: "Variance calculations between individual and colony level values"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VarianceCalculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 80
    canonical: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE
)
```
# Introduction of the problem
As seen in the quantitative genetics vignette, in SIMplyBee, traits are defined
and simulated at the individual (queen, worker, drone) level and mapped to
colony-level values. This requires specification of individual-level variances
and covariances for the queen and workers effects. However, in practice and
literature, we can obtain only colony-level values, and we would like to be able
to use these to inform the individual-level trait definitions. This requires
understanding how colony-level variances and covariances translate to
individual-level variances and covariances, as well as vice-versa.

This vignettes presents:
- the theory behind the relationship between individual-level and colony-level variances and covariances,
- the implementation of this theory in SIMplyBee, and
- examples of implementation for when the workers's values are combined as a mean or a sum.

# Theory: colony-level variance components
As seen in the quantitative genetics vignette, in SIMplyBee, traits are defined
and simulated at the individual (queen, worker, drone) level and mapped to
colony-level values. This requires specification of individual-level variances
and covariances for the queen and workers effects. However, in practice and
literature, we can obtain only colony-level values, and we would like to be able
to use these to inform the individual-level trait definitions. This requires
understanding how colony-level variances and covariances translate to
individual-level variances and covariances, as well as vice-versa.


We will follow the example from Brascamp and Bijma (2019):

$\sigma_{g_q}^2=0.5$ - this is their $\sigma_{A_q}^2=0.5$

$\sigma_{g_w}^2=1.0$ - this is their $\sigma_{A_w}^2=1.0$ among nominally
unrelated groups of workers

$\sigma_{\bar{g}_w}^2=0.32$ - this is their $\sigma_{\bar{A}_w}^2=0.32$ among
base related groups of workers

Where the relationship between the two is
$\sigma_{\bar{g}_w}^2 = a_{ii} * \sigma_{g_w}^2$ where $a_{ii}$ is average
additive genetic relationship between workers in a colony as can be defined (in
a non-inbred population) as
$a_{ii} = 1/4 + 1/2p_1 + 1/4(p_2 - p_1) + 1/4(1 - p_2)a_{ss}$; where $p_1$ is
the probability that two workers descend from the same drone, $p_2$ is the
probability that they descend from the same DPQ, and $a_{ss}$ is the relatedness
between two DPQs. We are ignore the $a_{ss}$ part.

$\sigma_{{g_q},{g_w}}=-0.35$

$\sigma_{P}^2 = \sigma_{\bar{g}_w^w}^2 + \sigma_{g_q^q}^2 + 2*\sigma_{\bar{g}_w^w g_q^q} + \sigma_{E}^2$

$\sigma_e^2=2.0$

$h^2=(a_{base}\sigma_{g,w}^2 + \sigma_{{g,w},{g,q}} + \sigma_{g,q}^2)/\sigma_{p,c}^2=???$

$h_w^2=a_{base}\sigma_{g,w}^2/\sigma_{p,c}^2=0.13$

$h_q^2=\sigma_{g,q}^2/\sigma_{p,c}^2=0.20$

$T^2=(\sigma_{g,w}^2 + 2\sigma_{{g,w},{g,q}} \sigma_{g,q}^2)/\sigma_{p,c}^2=0.32$

$T_w^2=\sigma_{g,w}^2/\sigma_{p,c}^2=0.41$

$T_q^2=\sigma_{g,q}^2/\sigma_{p,c}^2=0.20$

$r_g = cor(g_w, g_q) = -0.50$

TODO: show sensitivity to the number of workers - the larger the number of
workers, the smaller the variance between workers

$covA[1, 1] + nW * covA[2, 2] + 2 * k * covA[1, 2]$

Let's run again the example from the quantitative genetics vignette to have some
data to work with. In this example, we assumed that the contribution of workers
was summed up to obtain the combined colony-level worker contribution.
Accordingly, the variance for the worker trait was naively scaled by
$1/nWorkers$, which assumed that workers' contributions were independent. What
does this give us?

```{r load_variance_scripts, include = FALSE}
# load shared variance helper functions for this vignette
library("SIMplyBee")
library("dplyr")
library("ggplot2")
```

```{r FirstExampleFromQuanGenVignette}
library(package = "SIMplyBee")
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
nQtlPerChr <- 100
mean <- c(10, 10 / SP$nWorkers)
varA <- c(1, 1 / SP$nWorkers)
corA <- matrix(data = c( 1.0, -0.5, 
                        -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
             name = c("queenTrait", "workersTrait"))
varE <- c(3, 3 / SP$nWorkers)
corE <- matrix(data = c(3.0, 0.3, 
                        0.3, 3.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)
colonyGv <- calcColonyGv(apiary)
colonyPheno <- calcColonyPheno(apiary)

(popVar(colonyGv))
(popVar(colonyPheno))
```

In this example, we assumed that phenotypic variance on the individual level is
about 1/4 due to genetic variation and about 3/4 due to environmental variation,
but looking at colony genetic and phenotypic variances the ratio was about 1/2!

Also, if we naively compute the expected colony-level variances based on our assumptions of worker-independence when
setting individual-level variances, we get: 
```{r variance_expectation}
(covA <- corA * outer(X = sqrt(varA), Y = sqrt(varA), FUN = "*"))

# Trait environmental variation - on a per honeybee level
(covE <- corE * outer(X = sqrt(varE), Y = sqrt(varE), FUN = "*"))

# Trait phenotypic variation - on a per honeybee level
(covP <- covA + covE)

(covA[1, 1] + SP$nWorkers * covA[2, 2] + 2 * covA[1, 2])
(covP[1, 1] + SP$nWorkers * covP[2, 2] + 2 * covP[1, 2])
```
Here, we see an even larger discrepancy between the assumed and realised colony-level genetic variance,
where the realised values greatly exceed the assumed values.

So, let's have a look at how we can improve the simulation to get the desired
colony-level variances and covariances.

# Theory: Breaking down colony-level variance components

For genetic value of a colony $g_c$ we add up the queen's genetic value for the
queen effect $g_{q,q}$ and workers' genetic values for the workers effect
$\Sigma_{i=1}^{n_w}(g_{i,w})$:

$g_c = g_{q,q} + \Sigma_{i=1}^{n_w}(g_{i,w}).$

Expectation of colony genetic value is then:

$E(g_c) = E(g_{q,q} + \Sigma_{i=1}^{n_w}(g_{i,w}))$

$E(g_c) = E(g_{q,q}) + E(\Sigma_{i=1}^{n_w}(g_{i,w}))$

$E(g_c) = \mu_{g_{q}} + n_w E(g_{w,w})$

$E(g_c) = \mu_{g_{q}} + n_w \mu_{g_{w}}$

we are assuming here that $n_w$ is a fixed value, but if it is a random variable
then we would have

$E(g_c) = \mu_{g_{q}} + E(n_w) E(g_{w,w})$ using
<https://en.wikipedia.org/wiki/Distribution_of_the_product_of_two_random_variables>

$E(g_c) = \mu_{g_{q}} + \lambda_{n_w} \mu_{g_{w}}$

where $\lambda_{n_w}$ is average number of workers. So, in our case this turns
out the same as above. So in our case we have:

```{r}
nW <- SP$nWorkers
mean[1] + nW * mean[2]
mean(colonyGv)
mean(colonyPheno)
```

Variance of colony genetic value is then:

$Var(g_c) = Var(g_{q,q} + \Sigma_{i=1}^{n_w}(g_{i,w}))$

$Var(g_c) = Var(g_{q,q}) + Var(\Sigma_{i=1}^{n_w}(g_{i,w})) + 2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w}))$

$Var(g_c) = \sigma^2_{g_{q}} + Var(\Sigma_{i=1}^{n_w}(g_{i,w})) + 2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w}))$

So, we have three parts:

A)  the variance of the queen effect in queens $\sigma^2_{g_{q}}$,

B)  variance of a sum of workers effect in workers
    $Var(\Sigma_{i=1}^{n_w}(g_{i,w}))$, and

C)  covariance between the two $2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w}))$.

# A) Variance of the queen effect in queens $\sigma^2_{g_{q}}$ - in our case this is:

```{r check_queen_variances}
varA[1]
x = getQueenGv(apiary, collapse = TRUE)[, "queenTrait"]
var(x)
# ... note that R's var() divides by n-1, which matters with small n
popVar(x)
sum((x - mean[1])^2) / nColonies(apiary)

```

Anyway, quite close!

# B) Variance of a sum of workers effect in workers $Var(\Sigma_{i=1}^{n_w}(g_{i,w}))$

Given that this is a sum, we need to look at:

B1) variance of each of worker genetic values $Var(g_{i,w})$ and

B2) covariance between each pair of the values $Cov(g_{i,w}, g_{j,w})$.

This means, that family structure will start to matter because it induces
covariance between family members (workers). The simplest case is when all
workers come from the same father. Then we have:

$g_{1,w} = \frac{1}{2}g_{q,w} + g_{f,w} + r_{1,w}$

$g_{2,w} = \frac{1}{2}g_{q,w} + g_{f,w} + r_{2,w}$

...

$g_{n_w,w} = \frac{1}{2}g_{q,w} + g_{f,w} + r_{n_w,w}$

## B1) variance of each of worker genetic values $Var(g_{i,w})$

As stated initially, we have $Var(g_{i,w}) = \sigma^2_{g_{w}}$ and we have $n_w$
such terms, $n_w \sigma^2_{g_{w}}$. In our case this would be:

```{r check_workers_variances}
nW * varA[2]
g <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = "workersTrait")
var(g)
# ... note that R's var() divides by n-1, which matters with small n
sum((g - nW * mean[2])^2) / nColonies(apiary)
popVar(g)
```

So, a huge discrepancy between the variance of a sum of genetic values for the
workers effect `nW * varA[2]` (assuming independence) and a realised sum of
genetic values for the workers effect. As we have seen above, this is due to
family structure and associated B2) covariance between each pair of the values
$Cov(g_{i,w}, g_{j,w})$.

## B2) covariance between each pair of the values $Cov(g_{i,w}, g_{j,w})$

Clearly, these covariances matter a lot - there is lots of worker pairs in a
colony! How do these covariances look like?

$Cov(g_{1,w}, g_{2,w}) = Cov(\frac{1}{2}g_{q,w} + g_{f,w}, \frac{1}{2}g_{q,w} + g_{f,w})$

$Cov(g_{1,w}, g_{2,w}) = Var(\frac{1}{2}g_{q,w}) + Var(g_{f,w})$

assuming that mother and father are not related; further genetic variance
between drones is in fact half the genetic variance between queens because they
are haploid, so we get:

$Cov(g_{1,w}, g_{2,w}) = \frac{1}{4}\sigma^2_{g_w} + \frac{1}{4}\sigma^2_{g_w}$

$Cov(g_{1,w}, g_{2,w}) = \frac{1}{2}\sigma^2_{g_w}$.

Every pair of super-sister workers adds $\frac{1}{2}\sigma^2_{g_w}$ to B - note
that pair A-B adds this value, but so does the pair B-A, hence total is
$2\frac{1}{2}\sigma^2_{g_w} = \sigma^2_{g_w}$. With $n_w$ workers we get
$n_w n_w$ pairs (including with itself) or $n_w n_w - n_w$ pairs between
different workers. The total covariance contribution is then
$(n_w n_w - n_w)\frac{1}{2}\sigma^2_{g_w}$). In our case this would be:

```{r check_workers_variances2}
(varSumWorkers <- nW * varA[2]) # B1
(covSumWorkers <- (nW * nW - nW) * 1/2 * varA[2]) # B2
varSumWorkers + covSumWorkers # B

g <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = "workersTrait")
var(g) # B
# ... note that R's var() divides by n-1, which matters with small n
sum((g - nW * mean[2])^2) / nColonies(apiary) # B
popVar(g)
```

If we would only have super-sisters we would have \~1 for B1, \~49.5 for B2, and
\~50.5 for B, but with a mix of half-sisters and super-sisters we would have a
lower value, which is what we see above, \~26.8 for B.

TODO: Define the setting above in intro or start of M&M - We also can have
full-sisters, where the mother is obviously the same, but the fathers are
different, yet they come from the same mother, which is equivalent to a
full-sibs case in diploid species. We will assume in this work that we have
unrelated queens and unrelated drones (=base pop) - our aim is to set variances
in such a base population - to get the scale or the variances right. In
simulation of our base population we generated drones from virgin queens, which
means that our drones can in fact be brothers, so our realised variances might
be a bit off compared to this theory. But this is fine, we are mostly trying to
get the order of variances correct.

So, now we need to work out B2 with a mix of half-sisters and super-sisters :(
Sister workers are related only due to having the same mother, assuming that
drones the queen mated with are unrelated:

$g_{1,w} = \frac{1}{2}g_{q,w} + g_{f1,w} + r_{1,w}$

$g_{2,w} = \frac{1}{2}g_{q,w} + g_{f2,w} + r_{2,w}$

$Cov(g_{1,w}, g_{2,w}) = Cov(\frac{1}{2}g_{q,w} + g_{f1,w}, \frac{1}{2}g_{q,w} + g_{f2,w})$

$Cov(g_{1,w}, g_{2,w}) = Var(\frac{1}{2}g_{q,w})$

$Cov(g_{1,w}, g_{2,w}) = \frac{1}{4}Var(g_{q,w})$

Hence every pair of half-sister workers adds $\frac{1}{4}\sigma^2_{g_w}$ to B
(note, this is $2\frac{1}{4}\sigma^2_{g_w}$ for A-B and B-A pairs!). With $n_w$
workers and $n_f$ fathers we have $\frac{n_w}{n_f}$ workers per father. Further,
we have $n_f$ groups of super-sisters and $n_f n_f - n_f$ pairs of sister
groups. Assuming that half-sister groups are the same size, we have
$n_f \frac{n_w}{n_f} \frac{n_w}{n_f} = n_w \frac{n_w}{n_f} = \frac{n^2_w}{n_f}$
pairs of super-sisters and $(n_f n_f - n_f) \frac{n_w}{n_f} \frac{n_w}{n_f}$ -
this is
$n_f (n_f - 1) \frac{n_w}{n_f} \frac{n_w}{n_f} = (n_f - 1) n_w \frac{n_w}{n_f} = \frac{ (n_f - 1) n^2_w}{n_f}$.
Is this correct?

```{r pairs_of_workers}
(nF <- nFathers(colony))
nW * nW
(a <- nW * nW / nF)
(b <- (nF - 1) * nW * nW / nF)
a+b 
#TODO define nDPQ
```

Looks like!

So, super-sister pairs add $\frac{n^2_w}{n_f} \frac{1}{2}\sigma^2_{g_w}$ and
half-sister pairs add $\frac{ (n_f - 1) n^2_w}{n_f} \frac{1}{4}\sigma^2_{g_w}$.
Let's test this:

```{r check_workers_variances3}
(varSumWorkers <- nW * varA[2]) # B1
(covSumWorkersSuperSisters <- (nW * nW / nF) * 0.75 * varA[2]) # B2 in super-sisters ERROR: SHOULD BE 0.75
(covSumWorkersHalfSisters <- ((nF - 1) * nW * nW / nF) * 0.25 * varA[2]) # B2 in half-sisters
varSumWorkers + covSumWorkersSuperSisters + covSumWorkersHalfSisters # B

g <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = "workersTrait")
var(g) # B
# ... note that R's var() divides by n-1, which matters with small n
sum((g - SP$nWorkers * mean[2])^2) / nColonies(apiary) # B
```

Cool - very close!!! There can be some difference because the formulae above
assume that we have fixed sizes of half-sister and super-sister groups. Also,
there can be full-sisters in there as well!!!

TODO: We could also add full-sisters: To do this, we would have to assume how
many brother drones we use to figure out the grups

TODO: What if $n_w$ is a random variable? Following
<https://en.wikipedia.org/wiki/Distribution_of_the_product_of_two_random_variables>
we would use:

$Var(XY) = (Var(X) + E(X)^2) (Var(Y) + E(Y)^2) - E(X)^2 E(Y)^2$

but this is for two variables, while I have $n_w$ genetic values, so this
reference result is not that useful in our case! Well, if $n_w$ varies between
colonies, this must boost variance of colony-level genetic values significantly,
because kind of start calculating variance between apples (small colonies) and
oranges (large colonies).

# C) Covariance between the two $2Cov(g_{q,q}, \Sigma_{i-1}^{n_w}(g_{i,w}))$.

Let's repeat:

$g_{1,w} = \frac{1}{2}g_{q,w} + g_{f1,w} + r_{1,w}$

$g_{2,w} = \frac{1}{2}g_{q,w} + g_{f2,w} + r_{2,w}$ ...

The essential bit here is:

$Cov(g_{q,q}, g_{i,w}) = Cov(g_{q,q}, \frac{1}{2} g_{q,w})$

assuming that the queen and fathers are unrelated. Then:

$Cov(g_{q,q}, g_{i,w}) = Cov(g_{q}, \frac{1}{2} g_{w})$

$Cov(g_{q,q}, g_{i,w}) = \frac{1}{2} Cov(g_{q}, g_{w})$

$Cov(g_{q,q}, g_{i,w}) = \frac{1}{2} \sigma_{g_{q},g_{w}}$

So:

$2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w})) = 2Cov(g_{q,q}, n_w \frac{1}{2} g_{q,w})$

$2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w})) = 2 n_w \frac{1}{2}Cov(g_{q,q}, g_{q,w})$

$2Cov(g_{q,q}, \Sigma_{i=1}^{n_w}(g_{i,w})) = n_w \sigma_{g_{q},g_{w}}$

```{r check_workers_variances4}
nDPQ = 5
n_SS = (nW * nW / nF) - nW
n_FS = (nW * nW / nDPQ) -  (nW * nW / nF)
n_HS = (nW * nW / nDPQ) * (nDPQ - 1)

(n_SS + n_FS + n_HS + nW) == nW * nW

print("Genetic variances")
print("Part A")
(varQueen_A <- varA[1]) # A

print("Part B1")
(varSumWorkers_A <- nW * varA[2]) # B1
print("Part B2 in super-sisters")
(covSumWorkersSuperSisters_A <- n_SS * 0.75 * varA[2]) # B2 in super-sisters
print("Part B2 in full-sisters")
(covSumWorkersFullSisters_A <- n_FS * 0.50 * varA[2]) # B2 in full-sisters
print("Part B2 in half-sisters")
(covSumWorkersHalfSisters_A <- n_HS * 0.25 * varA[2]) # B2 in half-sisters
print("Part B combined")
(varCovSumWorkers_A <- varSumWorkers_A + covSumWorkersSuperSisters_A + covSumWorkersFullSisters_A + covSumWorkersHalfSisters_A) # B
print("Part C")
(covQueenSumWorkers_A <- nW * covA[1, 2]) # C

print("Part A + B + C")
varQueen_A + varCovSumWorkers_A + 2*covQueenSumWorkers_A

print("ColonyGv variance")
g <- colonyGv
var(g)
popVar(g)
sum((g - (mean(g)))^2) / nColonies(apiary)
```

Bingo for genetic variance.

```{r}


print("Environmental variance")
print("Part A")
(varQueen_E <- varE[1]) # A
print("Part B1")
(varSumWorkers_E <- nW * varE[2]) # B1, there is not environmental covariance between workers, so no B2
print("Part C")
(covQueenSumWorkers_E <- nW * covE[1, 2]) # C - both genetic and env cove here? TODO

print("Part A + B + C f")
varQueen_A + varCovSumWorkers_A + covQueenSumWorkers_A + varQueen_E + varSumWorkers_E + covQueenSumWorkers_E

p <- colonyPheno
var(p)
popVar(p)
sum((p - (mean(p)))^2) / nColonies(apiary)
```

And very close for phenotypic too, but we need to check some part above!

# SIMplyBee implementation

We have developed a set of functions that use the theory explained above to:
(i) calculate expected colony-level variances
and covariances based on the provided individual-level variances and covariances: `mapIndToColonyVar()`
(ii) calculate the required individual-level variances and covariances to achieve the desired 
colony-level variances and covariances: `mapColonyToIndVar()`

# Example: summarising worker effects as a sum

Let's see how these work in practice. We will use the same example as above, so let's just recall what values we've set.
Note again, that we are summarising the workers' effects as a sum, so we naively set the individual-level variance for the worker effect 
to be $1/nWorkers$ of the desired colony-level variance for the worker effect. 

Let's start off with a fresh simulation.

```{r}
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
nQtlPerChr <- 100
mean <- c(10, 10 / SP$nWorkers)
varA <- c(1, 1 / SP$nWorkers)
corA <- matrix(data = c( 1.0, -0.5, 
                        -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
             name = c("queenTrait", "workersTrait"))
varE <- c(3, 3 / SP$nWorkers)
corE <- matrix(data = c(3.0, 0.3, 
                        0.3, 3.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)
```

Now, let's compute the corresponding colony-level variances and covariances using the `mapIndToColonyVar()` function.
```{r mapIndToColonyVar_sum}
# When the function to summarise worker effects is SUM
calcColonyValues <- mapIndToColonyVar(varA_q = varA[1], 
                  varA_w= varA[2],
                  corA_qw = corA[1, 2],
                  varE_q = varE[1],
                  varE_w = varE[2],
                  corE_qw = corE[1, 2],
                  nW = nW, nF = nF,
                  nDPQ = nDPQ,
                  workersFUN = "sum")

calcColonyValues$varA_c
colonyGv <- calcColonyGv(apiary, FUN = mapCasteToColonyGv, workersFUN = colSums)
var(colonyGv)
```
We see that the realised and calculated colony-level variances match quite closely. 

Now, let's do it the other way around. This is a more realistic scenario since you would normally obtain 
the colony-level variances and covariances from the literature and would want to recreate them using simulation.
Here, we assume, you obtain a separate estimate for the queen and for the worker trait.
Let's say we want to achieve the following colony-level variances and covariances:
```{r mapColonyToIndVar_sum}
varA_q <- 1
varA_wbar <- 1
corA_qwbar <- -0.5

varE_q <- 3
varE_wbar <- 3
corE_qwbar <- 0.3

nW <- 100
nF <- 15
nDPQ <- 5

calcIndValues <- mapColonyToIndVar(varA_q = varA_q, varA_wbar = varA_wbar, corA_qwbar = corA_qwbar, 
                                varE_q = varE_q, varE_wbar = varE_wbar, corE_qwbar = corE_qwbar, 
                                nW = nW, nF = nF, nDPQ = nDPQ, workersFUN = "sum")
```
Now, let's use these calculated individual-level variances and covariances to simulate a new population and check if we achieve the desired colony-level variances and covariances.


```{r example_sum}
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
SP$nWorkers = nW
SP$nFathers = nF
nQtlPerChr <- 100
mean <- c(0, 0)
varA <- c(calcIndValues$varA_q, calcIndValues$varA_w)
corA <- matrix(data = c( 1.0, calcIndValues$corA_qw,
                        calcIndValues$corA_qw,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
            name = c("queenTrait", "workersTrait"))
varE <- c(calcIndValues$varE_q, calcIndValues$varE_w)
corE <- matrix(data = c(1.0, calcIndValues$corE_qw,
                        calcIndValues$corE_qw, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)

drones <- createDrones(x = basePop[1:nDPQ], nInd = 3)
colony <- createColony(x = basePop[nDPQ+1])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:nDPQ], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)

# Get the realised numbers
real_A_q <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = 1, workersTrait = NULL)
real_varA_q <- popVar(real_A_q)
real_A_wbar <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = 2, workersFUN = colSums)
real_varA_wbar <- popVar(real_A_wbar)
real_A_c <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = 1, workersTrait = 2, workersFUN = colSums)
real_varA_c <- popVar(real_A_c)

real_P_q <- calcColonyPheno(apiary, mapCasteToColonyPheno, queenTrait = 1, workersTrait = NULL)
real_E_q <- real_P_q - real_A_q
real_varE_q <- popVar(real_E_q)
real_P_wbar <- calcColonyPheno(apiary, mapCasteToColonyPheno, queenTrait = NULL, workersTrait = 2, workersFUN = colSums)
real_E_wbar <- real_P_wbar - real_A_wbar
real_varE_wbar <- popVar(real_E_wbar)
real_P_c <- calcColonyPheno(apiary, mapCasteToColonyPheno, queenTrait = 1, workersTrait = 2, workersFUN = colSums)
real_E_c <- real_P_c - real_A_c
real_varE_c <- popVar(real_E_c)

variances <- data.frame(set_varA_q = varA_q, real_varA_q = real_varA_q[1,1],
                        set_varA_wbar = varA_wbar, real_varA_wbar = real_varA_wbar[1,1],
                        real_varA_c = real_varA_c[1,1], 
                        set_varE_q = varE_q, real_varE_q = real_varE_q[1,1],
                        set_varE_wbar = varE_wbar, real_varE_wbar = real_varE_wbar[1,1],
                        real_varE_c = real_varE_c[1,1],
                        fun = "sum")

# Plot the results
variancesLong <- variances %>%
  tidyr::pivot_longer(cols = -c(fun), names_to = "VarType", values_to = "Var")
variancesLong$RealCalc <- sapply(strsplit(as.character(variancesLong$VarType), "_"), "[", 1)
# Extract the second and third elements and paste them together to get the variance type for plotting
variancesLong$VarTypePlot <- paste0(sapply(strsplit(as.character(variancesLong$VarType), "_"), "[", 2), "_", sapply(strsplit(as.character(variancesLong$VarType), "_"), "[", 3))

variancesLong$VarTypePlot <- factor(variancesLong$VarTypePlot, levels = c("varA_q", "varA_wbar", "varA_c",
                                                                        "varE_q", "varE_wbar", "varE_c"))

variancesLong |>   filter(VarTypePlot %in% c("varA_q", "varA_wbar", "varE_q", "varE_wbar")) |>
  ggplot(aes(x = VarTypePlot, y = Var, fill = RealCalc)) +
  geom_col(position = "dodge") +
  facet_wrap(. ~fun, scales = "free") +
  theme_bw(base_size = 20) + 
  scale_fill_manual(values = c("#D55E00", "#084159"))  # one color per RealCalc level (set/real)
```

As we see, we are pretty much on point with matching the desired values! There is of course the element of stochasticity and
the limited samples size. 


# Example: summarising worker effects as a mean
Let's repeat all of the above using a mean to summarise workers' values.
For this, we set the individual-level variance for the worker effect the same as the desired colony-level variance.
For this, we need to resimulate our apiary.

```{r simulation_mean}
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
nQtlPerChr <- 100
mean <- c(10, 10)
varA <- c(1, 1)
corA <- matrix(data = c( 1.0, -0.5, 
                        -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
             name = c("queenTrait", "workersTrait"))
varE <- c(3, 3 / SP$nWorkers)
corE <- matrix(data = c(3.0, 0.3, 
                        0.3, 3.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)
```

Let's again compute the corresponding colony-level variances and covariances using the `mapIndToColonyVar()` function.
```{r mapIndToColonyVar_mean}
# When the function to summarise worker effects is MEAN
calcColonyValues <- mapIndToColonyVar(varA_q = varA[1], 
                  varA_w= varA[2],
                  corA_qw = corA[1, 2],
                  varE_q = varE[1],
                  varE_w = varE[2],
                  corE_qw = corE[1, 2],
                  nW = nW, nF = nF,
                  nDPQ = nDPQ,
                  workersFUN = "mean")

calcColonyValues$varA_c
colonyGv <- calcColonyGv(apiary, FUN = mapCasteToColonyGv, workersFUN = colMeans)
var(colonyGv)
```
We again see that the realised and calculated colony-level variances match very closely. 

And for the other way around:
```{r mapColonyToIndVar_mean}
calcIndValues <- mapColonyToIndVar(varA_q = varA_q, varA_wbar = varA_wbar, corA_qwbar = corA_qwbar, 
                                varE_q = varE_q, varE_wbar = varE_wbar, corE_qwbar = corE_qwbar, 
                                nW = nW, nF = nF, nDPQ = nDPQ, workersFUN = "sum")
```
Now, let's use these calculated individual-level variances and covariances to simulate a new population and check if we achieve the desired colony-level variances and covariances.


```{r example_mean}
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
SP$nWorkers = nW
SP$nFathers = nF
nQtlPerChr <- 100
mean <- c(0, 0)
varA <- c(calcIndValues$varA_q, calcIndValues$varA_w)
corA <- matrix(data = c( 1.0, calcIndValues$corA_qw,
                        calcIndValues$corA_qw,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
            name = c("queenTrait", "workersTrait"))
varE <- c(calcIndValues$varE_q, calcIndValues$varE_w)
corE <- matrix(data = c(1.0, calcIndValues$corE_qw,
                        calcIndValues$corE_qw, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)

drones <- createDrones(x = basePop[1:nDPQ], nInd = 3)
colony <- createColony(x = basePop[nDPQ+1])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:nDPQ], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)

# Get the realised numbers
real_A_q <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = 1, workersTrait = NULL)
real_varA_q <- popVar(real_A_q)
real_A_wbar <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersTrait = 2, workersFUN = colSums)
real_varA_wbar <- popVar(real_A_wbar)
real_A_c <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = 1, workersTrait = 2, workersFUN = colSums)
real_varA_c <- popVar(real_A_c)

real_P_q <- calcColonyPheno(apiary, mapCasteToColonyPheno, queenTrait = 1, workersTrait = NULL)
real_E_q <- real_P_q - real_A_q
real_varE_q <- popVar(real_E_q)
real_P_wbar <- calcColonyPheno(apiary, mapCasteToColonyPheno, queenTrait = NULL, workersTrait = 2, workersFUN = colSums)
real_E_wbar <- real_P_wbar - real_A_wbar
real_varE_wbar <- popVar(real_E_wbar)
real_P_c <- calcColonyPheno(apiary, mapCasteToColonyPheno, queenTrait = 1, workersTrait = 2, workersFUN = colSums)
real_E_c <- real_P_c - real_A_c
real_varE_c <- popVar(real_E_c)

variances <- data.frame(set_varA_q = varA_q, real_varA_q = real_A_q[1,1],
                        set_varA_wbar = varA_wbar, real_varA_wbar = real_varA_wbar[1,1],
                        real_varA_c = real_varA_c[1,1], 
                        set_varE_q = varE_q, real_varE_q = real_varE_q[1,1],
                        set_varE_wbar = varE_wbar, real_varE_wbar = real_varE_wbar[1,1],
                        real_varE_c = real_varE_c[1,1],
                        fun = "sum")

# Plot the results
variancesLong <- variances %>%
  tidyr::pivot_longer(cols = -c(fun), names_to = "VarType", values_to = "Var")
variancesLong$RealCalc <- sapply(strsplit(as.character(variancesLong$VarType), "_"), "[", 1)
# Extract the second and third elements and paste them together to get the variance type for plotting
variancesLong$VarTypePlot <- paste0(sapply(strsplit(as.character(variancesLong$VarType), "_"), "[", 2), "_", sapply(strsplit(as.character(variancesLong$VarType), "_"), "[", 3))

variancesLong$VarTypePlot <- factor(variancesLong$VarTypePlot, levels = c("varA_q", "varA_wbar", "varA_c",
                                                                        "varE_q", "varE_wbar", "varE_c"))

variancesLong |>   filter(VarTypePlot %in% c("varA_q", "varA_wbar", "varE_q", "varE_wbar")) |>
  ggplot(aes(x = VarTypePlot, y = Var, fill = RealCalc)) +
  geom_col(position = "dodge") +
  facet_wrap(. ~fun, scales = "free") +
  theme_bw(base_size = 20) + 
  scale_fill_manual(values = c("#D55E00", "#084159"))  # one color per RealCalc level (set/real)
```

