---
title: "Variance calculations between individual and colony level values"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VarianceCalculations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 80
    canonical: true
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  include = TRUE
)
```

```{r cor2cov, include = FALSE}
library(package = "SIMplyBee")
cor2cov <- function(corr, var) {
  # Ensure inputs are matrices/vectors
  corr <- as.matrix(corr)
  var  <- as.numeric(var)

  # Standard deviations
  sd <- sqrt(var)

  # Covariance matrix
  cov <- corr * (sd %o% sd)

  return(cov)
}
```

Let's again run the example from the quantitative genetic vignette.

```{r FirstExampleFromQuanGenVignette}
# Run the example from the quantitative genetic vignette
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
nQtlPerChr <- 100
mean <- c(10, 10 / SP$nWorkers)
varA <- c(1, 1 / SP$nWorkers)
corA <- matrix(data = c( 1.0, -0.5,
                         -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
             name = c("queenTrait", "workersTrait"))
varE <- c(3, 3 / SP$nWorkers)
corE <- matrix(data = c(1.0, 0.3,
                        0.3, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)
head(basePop@gv)
head(basePop@pheno)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)
colonyGv <- calcColonyGv(apiary)
colonyPheno <- calcColonyPheno(apiary)

```

How the expected and realised variances look like
```{r ExpectedAndRealisedVariances}
corA
(covA <- corA * outer(X = sqrt(varA), Y = sqrt(varA), FUN = "*"))

# Trait environmental variation - on a per honeybee level
(covE <- corE * outer(X = sqrt(varE), Y = sqrt(varE), FUN = "*"))

# Trait phenotypic variation - on a per honeybee level
(covP <- covA + covE)
(corP <- cov2cor(covP))
(varP <- diag(covP))
nW <- SP$nWorkers


k <- 1
print("Expected genetic and phenotypic variance")
covA[1, 1] + nW * covA[2, 2] + 2 * k * covA[1, 2] # variance can not be negative!
covP[1, 1] + nW * covP[2, 2] + 2 * k * covP[1, 2]

print("Realised genetic and phenotypic variance")
var(colonyGv)
var(colonyPheno)
```

Now, in theory, how we go from the individual to colony level variances?
The queen trait variance remains the same, since there is only one queen. The worker
individual variance needs to be adjusted due to covariances between workers.

```{r AdjustingWorkerVariance}
a_ii = 0.40
varA_colony <- c(varA[1], varA[2] * a_ii)
corA = cov2cor(covA)
covA_colony = cor2cov(corA, var = varA_colony)

covA[1, 1] + nW * covA[2, 2] + 2 * k * covA[1, 2]
covA_colony[1, 1] + covA_colony[2, 2] + 2 * k * covA_colony[1, 2]

var(colonyGv)
var(colonyPheno)
```

We need to add the covariance due to different types of workers in the colony
```{r CovariancesBetweenWorkers}
mapIndToColonyVar <- function(varA_q, varA_w, corA_qw, nW, nF, nDPQ, workersFUN = "sum") {
  if (workersFUN == "mean") {
    factor = 1 / nW^2
  } else if (workersFUN == "sum") {
    factor = nW
  }

  # Determine how many pairs of each you have
  nW = nW
  n_SS = (nW * nW / nF) - nW
  n_FS = (nW * nW / nDPQ) -  (nW * nW / nF)
  n_HS = (nW * nW / nDPQ) * (nDPQ - 1)

  A <- varA_q
  if (workersFUN == "mean") {
    B1 = 1 / nW * varA_w
  } else if (workersFUN == "sum") {
    B1 = nW * varA_w
  }

  B2_ss <- n_SS * 0.75 * varA_w
  B2_fs <- n_FS * 0.50 * varA_w
  B2_hs <- n_HS * 0.25 * varA_w
  if (workersFUN == "mean") {
    B = B1 + 1/nW^2 * (B2_ss + B2_fs + B2_hs)
  } else if (workersFUN == "sum") {
    B = B1 + B2_ss + B2_fs + B2_hs
  }

  covA_qw = corA_qw * sqrt(varA_q) * sqrt(varA_w)

  if (workersFUN == "mean") {
    C <- covA_qw
  } else if (workersFUN == "sum") {
    C <- nW * covA_qw
  }
  
  varC <- A + B + C

  return(list(A = A, B = B, C = C, colonyVar = varC))
}

#This is the same as above, just simplified
# mapIndToColonyVar <- function(varA_q, varA_w, corA_qw, nW, nF, nDPQ, workersFUN = "sum") {

#   # Determine how many pairs of each you have
#   p_SS = 1 / nF - 1 / nW
#   p_FS = (1 / nDPQ) -  (1 / nF)
#   p_HS = (1 / nDPQ) * (nDPQ - 1)
#   A <- varA_q
#   B1 <- nW * varA_w
#   B2_ss <- p_SS * 0.75
#   B2_fs <- p_FS * 0.50
#   B2_hs <- p_HS * 0.25
  
#   covA_qw = corA_qw * sqrt(varA_q) * sqrt(varA_w)
#   C <- nW * covA_qw
  
#   if (workersFUN == "sum") {
#     varC <- A + B1 + (B2_ss + B2_fs + B2_hs)*nW^2*varA_w + C
#   } else if (workersFUN == "mean") {
#     varC <- A + varA_w + (B2_ss + B2_fs + B2_hs)*varA_w + covA_qw# This is equal as varA_w * average relatedness within a colony
#   }
#   return(varC)
# }

```

```{r testingMapIndToColonyVar}
# Get the real number of fathers and DPQs
nW = round(mean(nWorkers(apiary)), 0)
nF = round(mean(nFathers(apiary)), 0)
nDPQ <- round(mean(sapply(getFathers(apiary), function(x) length(unique(x@mother)))), 0)


# When the function to summarise worker effects is SUM
mapIndToColonyVar(varA_q = varA[1], 
                  varA_w= varA[2],
                  corA_qw = corA[1,2],
                  nW = nW, nF = nF,
                  nDPQ = nDPQ,
                  workersFUN = "sum")
colonyGv <- calcColonyGv(apiary, FUN = mapCasteToColonyGv, workersFUN = colSums)
var(colonyGv)
```


```{r FirstExampleFromQuanGenVignette}
# Run the example from the quantitative genetic vignette
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
nQtlPerChr <- 100
mean <- c(10, 10)
varA <- c(1, 1)
corA <- matrix(data = c( 1.0, -0.5,
                         -0.5,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
             name = c("queenTrait", "workersTrait"))
varE <- c(3, 3)
corE <- matrix(data = c(1.0, 0.3,
                        0.3, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)
head(basePop@gv)
head(basePop@pheno)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)
colonyGv <- calcColonyGv(apiary)
colonyPheno <- calcColonyPheno(apiary)

```

```{r testingMapIndToColonyVarMean}
# Get the real number of fathers and DPQs
nW = round(mean(nWorkers(apiary)), 0)
nF = round(mean(nFathers(apiary)), 0)
nDPQ <- round(mean(sapply(getFathers(apiary), function(x) length(unique(x@mother)))), 0)


# When the function to summarise worker effects is MEAN
mapIndToColonyVar(varA_q = varA[1], 
                  varA_w= varA[2],
                  corA_qw = corA[1,2],
                  nW = nW, nF = nF,
                  nDPQ = nDPQ,
                  workersFUN = "mean")
colonyGv <- calcColonyGv(apiary, FUN = mapCasteToColonyGv, workersFUN = colMeans)
var(colonyGv)
```


```{r mapColonyToIndVar}
# First option, we know the varA_q, varA_wbar and cor, and one varE
mapColonyToIndVar <- function(varA_q,
                              varA_wbar,
                              corA_qwbar = NULL,
                              varE,
                              nW,
                              nF,
                              nDPQ,
                              workersFUN = "sum") {

  # scaling factor
  if (workersFUN == "sum") {
    covA_qwbar <- corA_qwbar * sqrt(varA_q) * sqrt(varA_wbar)
    covA_qw <- covA_qwbar / nW 
  } else if (workersFUN == "mean") {
    covA_qw <- covA_qwbar
  }

  # pair counts
  n_SS <- (nW * nW / nF) - nW
  n_FS <- (nW * nW / nDPQ) - (nW * nW / nF)
  n_HS <- (nW * nW / nDPQ) * (nDPQ - 1)

  # worker variance coefficient
  if (workersFUN == "sum") {
  K <- nW +
        n_SS * 0.75 +
        n_FS * 0.50 +
        n_HS * 0.25
  } else if (workersFUN == "mean") {
    K <- 1 / nW + ((n_SS * 0.75 + 
                    n_FS * 0.50 + 
                    n_HS * 0.25) / nW^2)
  }

  varA_w <- varA_wbar / K

  corA_qw <- covA_qw / (sqrt(varA_q) * sqrt(varA_w))

  return(list(varA_q = varA_q, varA_wbar = varA_wbar, varA_w = varA_w, covA_qwbar = covA_qwbar, covA_qw = covA_qw, corA_qw = corA_qw))
}
```

```{r testMapColonyToIndFunction}
mapColonyToIndVar(varA_q = 10, varA_wbar = 10, corA_qwbar = -0.5, varE = 30, nW = 100, nF = 5, nDPQ = 5, workersFUN = "sum")

# Run the example from the quantitative genetic vignette
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
SP$nWorkers <- 100
nQtlPerChr <- 100
mean <- c(0, 0)
varA <- c(10, 0.002836879)
corA <- matrix(data = c( 1.0, -0.2968586,
                         -0.2968586,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
             name = c("queenTrait", "workersTrait"))
varE <- c(3, 3)
corE <- matrix(data = c(1.0, 0.3,
                        0.3, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)
head(basePop@gv)
head(basePop@pheno)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)

workerGroupGv <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersFUN = colSums)
colonyGv <- calcColonyGv(apiary) # sum is the default

print("Worker group variance")
var(workerGroupGv)
print("Colony variance")
var(colonyGv)
```

```{r testMapColonyToIndFunction}
mapColonyToIndVar(varA_q = 10, varA_wbar = 10, corA_qwbar = -0.5, varE = 30, nW = 100, nF = 5, nDPQ = 5, workersFUN = "mean")

# Run the example from the quantitative genetic vignette
founderGenomes <- quickHaplo(nInd = 20, nChr = 16, segSites = 1000)
SP <- SimParamBee$new(founderGenomes)
SP$nWorkers <- 100
nQtlPerChr <- 100
mean <- c(0, 0)
varA <- c(10, 28.36879)
corA <- matrix(data = c( 1.0, -0.2968586,
                         -0.2968586,  1.0), nrow = 2, byrow = TRUE)
SP$addTraitA(nQtlPerChr = nQtlPerChr, mean = mean, var = varA, corA = corA,
             name = c("queenTrait", "workersTrait"))
varE <- c(3, 3)
corE <- matrix(data = c(1.0, 0.3,
                        0.3, 1.0), nrow = 2, byrow = TRUE)
SP$setVarE(varE = varE, corE = corE)
basePop <- createVirginQueens(founderGenomes, n = 20)
head(basePop@gv)
head(basePop@pheno)
drones <- createDrones(x = basePop[1:5], nInd = 3)
colony <- createColony(x = basePop[6])
colony <- cross(x = colony, drones = drones, checkCross = "warning")
colony <- addWorkers(x = colony, nInd = 50)
colony <- buildUp(colony)
apiary <- createMultiColony(basePop[7:20])
drones <- createDrones(basePop[1:5], nInd = 100)
apiary <- cross(x = apiary, drones = drones, crossPlan = "create", checkCross = "warning")
apiary <- buildUp(apiary)

workerGroupGv <- calcColonyGv(apiary, mapCasteToColonyGv, queenTrait = NULL, workersFUN = colSums)
colonyGv <- calcColonyGv(apiary) # sum is the default

var(workerGroupGv)
var(colonyGv)
```